<!-- To be compatible with workflowr, we need to specify the absolute directory under which the plots are generated, otherwise they will be saved into figure/<basename>.Rmd/<chunk-name>.<extension. and not visible in the html document: https://github.com/workflowr/workflowr/issues/111 -->
```{r template-options, include = FALSE}
library(ggbeeswarm)
knitr::opts_knit$set(
  base.dir = "docs"
  )
```



#### Download results

<!-- Write DGE results to xlsx file -->
```{r write-dge-xlsx}
# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'
file_xlsx <- file.path('./docs/file',rmd_file, paste0('dge.',file_prefix,'.xlsx'))

dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)

# Generate workbook
wb <- createWorkbook()
for(i in names(dge_list)) {
  addWorksheet(wb, i)
  res <- dge_list[[i]]
  writeData(wb, i, res)
}
saveWorkbook(wb, file_xlsx, TRUE)

dge_file_xlsx <- file_xlsx
```


- [**Differential gene expression**](`r gsub("docs/", "" , dge_file_xlsx)`)

#### Summary of results
```{r dge-summary-fdr-conf}
use_names <- setdiff(names(dge_list), 'combined')
```


##### FDR distribution
```{r dge-summary-fdr-dist}
i <- use_names[1]
res <- foreach(i = use_names, .combine = rbind) %do% {
  x <- dge_list[[i]]
  c(
    `P-adjusted < 0.1` = x %>% filter(padj <= 0.1) %>% nrow,
    `P-adjusted < 0.05` = x %>% filter(padj <= 0.05) %>% nrow,
    `P-adjusted < 0.01` = x %>% filter(padj <= 0.01) %>% nrow,
    `P-adjusted < 0.001` = x %>% filter(padj <= 0.001) %>% nrow,
    `P-adjusted < 0.0001` = x %>% filter(padj <= 0.0001) %>% nrow
  )
}
if(class(res)[1] == 'integer'){
  res <- data.frame(res) %>% t() 
}

rownames(res) <-use_names
res %>% 
  data.frame(check.names = F) %>% 
  rownames_to_column('comparison') %>% 
  kbl(caption = 'Number of genes at different FDR thresholds') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  column_spec(1, bold = TRUE)
```

##### Number of genes at FDR ≤ 0.05
```{r dge-summary-fdr-05}
i <- use_names[1]
res <- foreach(i = use_names, .combine = rbind) %do% {
  x <- dge_list[[i]]%>% 
    filter(!is.na(padj)) %>%
    filter(padj <= p_thrs)
  c(
    `Up` = x %>% filter(log2FoldChange > 0) %>% nrow,
    `Down` = x %>% filter(log2FoldChange < -0) %>% nrow
  )
}
if(class(res)[1] == 'integer'){
  res <- data.frame(res) %>% t() 
}

rownames(res) <- use_names
res %>% 
  data.frame(check.names = F) %>% 
  rownames_to_column('comparison') %>% 
  kbl(caption = 'Number of genes at adjusted P value < 0.05') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  column_spec(1, bold = TRUE)
cat("\n\n")
```



#### Volcano-plots
Top-20 genes based on adjusted P-values are labeled.

```{r dge-volcano-plot, echo=FALSE, results='asis'}
i=use_names[1]
for(i in use_names) {
  cat("##### Comparison:", i, "\n\n")
  # cat(dge_list[[i]]$comparison, "\n\n")
  x <- dge_list[[i]]
  res <- plot.volcano(
    x$log2FoldChange, 
    x$padj, 
    label = x$gene_name, 
    fill.palette = c("gray80", "red"), 
    top.labels = 20, 
    pval.thr = p_thrs, 
    hline.color =  'grey60', hline.type = 2, hline.size = 0.5) +
    geom_vline(xintercept = 0, linetype = 2, linewidth = 0.5, color = 'grey60')
  
  subchunkify(print(res), 
                fig.width = 3.5, 
                fig.height = 3.5*0.618,
                dpi = 600,
                prefix = paste0('dge-volcano-', file_prefix),
                suffix = tolower(i) %>% gsub(" ", "-", .))
  
  cat("\n\n")
}
```


#### Heatmaps

```{r dge-heatmap-plot, echo=FALSE, results='asis'}
i <- use_names[1]
# p_thrs <- 0.05
# lfc_thrs <- 1
topn <- 50
max_color_range <- 3 # put the color range to -3 to 3

for(i in use_names) {
  cat("##### Comparison", i, "\n\n")

  # Extract samples from comparisons
  use_samples <- subset_list[[use_subset]]
  use_se <- se[,use_samples]
  
  # Arrange samples by condition (first control and then sample)
  use_annot <- colData(use_se) %>% data.frame %>%
    arrange(condition, ko)
    # mutate(
    #   condition_new = ifelse(sample_alias %in% use_comp[[1]], 
    #                          names(use_comp)[1], 
    #                          names(use_comp)[2]),
    #   condition_new = factor(condition_new, levels = names(use_comp)[2:1]),
    #   condition_num = as.numeric(condition_new)
    # ) %>% 
    # arrange(condition_num)
  use_samples <- rownames(use_annot)
  
  # Extract DGE results
  use_dge <- dge_list[[i]] %>% 
    filter(!is.na(padj)) %>%
    filter(padj < p_thrs & abs(log2FoldChange) > lfc_thrs) %>% 
    mutate(direction = ifelse(log2FoldChange > 0, 'Up', 'Down')) %>% 
    arrange(padj) %>% 
    head(topn) %>% 
    arrange(log2FoldChange)
  if(nrow(use_dge) > 50){
    cat(paste0("Showing ", nrow(use_dge)," features with FDR < ", p_thrs, " and absolute log2 fold-change > ", lfc_thrs, "\n"))
  } else {
    cat(paste0("Only ", nrow(use_dge)," features with FDR < ", p_thrs, " and absolute log2 fold-change > ", lfc_thrs, ". Showing top-", topn," features based on P value and absolute log2 fold-change >", lfc_thrs,"\n"))
    use_dge <- dge_list[[i]]  %>% 
      filter(!is.na(padj)) %>%
      filter(padj < p_thrs & abs(log2FoldChange) > lfc_thrs) %>% 
      mutate(direction = ifelse(log2FoldChange > 0, 'Up', 'Down')) %>% 
      arrange(padj) %>% 
      head(topn) %>% 
      arrange(log2FoldChange)
  }
  
  if(nrow(use_dge) > 2){
  # Extract matrix of expression
  mat <- assay(use_se, 'logcounts')[use_dge$gene_name, use_samples]
  # rownames(mat) <- make.names(use_dge$gene_name, unique = TRUE)
  zmat <- t(apply(mat, 1, scale, center = TRUE, scale = TRUE))
  colnames(zmat) <- colnames(mat)
  zmat <- zmat[,rownames(use_annot)]
  
  # Col Annotation
  use_annot_levels <-levels(use_annot$condition_new)
  column_ha <-  HeatmapAnnotation(
    condition = anno_block(
      # gp = gpar(fill = c('#fdbf6f', '#6a3d9a') %>% set_names(use_annot_levels)),
      labels = use_annot_levels, 
      labels_gp = gpar(col = "black", fontsize = 8)
    ),
    `Condition` = use_annot$condition,
    `KO` = use_annot$ko,
    col = list(
      `Condition` = condition_palette_red,
      `KO` = ko_palette
    )
  )
    
  # Row annotation (right)
  row_annot_right <-  rowAnnotation(
    log2FC = anno_numeric(
      round(use_dge$log2FoldChange,1), 
      labels_gp = gpar(fontsize = 6), 
      bg_gp = gpar(
        fill = c("#74add1", "#f46d43"), 
        col = 'white'
      ),
      bar_width = unit(1, "npc") - unit(1.5, "pt")
    ), 
    annotation_name_rot = 0)
  
  # Row annotation (left)
  direction_palette <- c('Up' = "#f46d43", 'Down' = "#74add1")
  direction_palette <- direction_palette[names(direction_palette) %in% use_dge$direction]
  
  row_annot_left <-  rowAnnotation(
    direction = anno_block(
      gp = gpar(fill = direction_palette),
      labels = names(direction_palette), 
      labels_gp = gpar(col = "white", fontsize = 8)
    )
  )
  
  # Heatmap color
  zmax <- zmat %>% abs %>% max
  if(!is.null(max_color_range))
    zmax <- max_color_range
  col_fun <-  colorRamp2(
    seq(-zmax, zmax, length.out = 11),
    rev(brewer.pal(11, "RdBu"))
  )
  
  # Generate Heatmap
  htz <- Heatmap(
    zmat,
    name= 'z-score',
    col = col_fun,
    row_names_gp = gpar(fontsize = 8),
    column_names_gp = gpar(fontsize = 8),
    
    top_annotation = column_ha,
    column_split = use_annot$condition_num,
    #cluster_column_slices = turn off the clustering on slices, and now you can precisely control the order of slices
    cluster_column_slices = FALSE, 
    column_title = NULL,
    
    right_annotation = row_annot_right,
    row_split = use_dge$direction,
    # left_annotation = row_annot_left,
    row_title = NULL
  )
  
  fig_height <- (7.2*0.3) + nrow(zmat)*0.15
  fig_height <- ifelse(fig_height < 7, 7, fig_height)
  
  subchunkify(
    draw(
      htz, 
      annotation_legend_side = "bottom"
    ), 
    fig.width = 7.2, 
    fig.height = fig_height,
    dpi = 600,
    prefix = paste0('dge-heatmap-plot-', file_prefix),
    suffix = tolower(i) %>% gsub(" ", "-", .)
  )
  cat("\n\n")
    
    
  } else {
    cat(paste0("Less than 2 genes with FDR < ", p_thrs, " and absolute log2 fold-change > ", lfc_thrs, "\n"))
  }
    
  cat("\n\n")
  
}
```


#### Fold-change plots

Plot comparing the fold-change of shear stress in KO versus WT conditions
```{r log2fc-scatter}
use_data <- dge_list[['combined']] %>%
  filter(!is.na(padj_wt) | !is.na(padj_ko)) %>%
  filter(!is.na(padj_wt) & !is.na(padj_ko)) %>%
  mutate(padj_int = ifelse(padj_int > p_limit, padj_int, p_limit)) %>%
  arrange(desc(padj_int))
```

Scatter plot labeling top-20 most significant genes based on adjusted P value for interaction
```{r log2fc-scatter-top20, echo=FALSE, results='asis'}
label_genes <- use_data %>% tail(20)

res <- use_data %>% 
  ggplot(aes(log2FoldChange_ko, log2FoldChange_wt,
             color = -log10(padj_int), size = -log10(padj_int))) +
  geom_abline(lty = 'dashed', color = 'gray60', linewidth = one_pt/2) +
  geom_hline(yintercept = 0, lty = 'dashed', color = 'gray60', linewidth = one_pt/2) +
  geom_vline(xintercept = 0, lty = 'dashed', color = 'gray60', linewidth = one_pt/2) +
  geom_point(pch = 16, alpha = 0.7) +
  geom_text_repel(aes(label = gene_name), 
                  data = label_genes, 
                  min.segment.length = 0, 
                  seed = 42, 
                  box.padding = 0.5, 
                  max.overlaps = Inf, 
                  show.legend = FALSE, 
                  size = geom_text_size,
                  segment.size = one_pt/2,
                  segment.alpha = 0.7) +
  scale_color_viridis_c(direction = -1) +
  scale_size_continuous(range = c(0.1, 2)) +
  labs(
    title = paste('Differential expression analysis in', use_subset),
    y = 'log2FC shear stress in wt',
    x = 'log2FC shear stress in ko',
    color = '-log10(padj)\ninteraction',
    caption = paste('Adjusted P value capped at', p_limit)
  ) #+
  # scale_colour_gradientn(
  #   colours = c("#E6E9AF", "#8EA3A6", "#605EA1", "#22177A"),
  #   rescaler = ~ scales::rescale_mid(.x, mid = -log10(0.0001))
  # )

  subchunkify(
    print(res), 
    fig.width = 3.6, 
    fig.height = 3.6*0.618,
    dpi = 600,
    prefix = paste0('log2fc-scatter-top20-', file_prefix)
    # suffix = tolower(i) %>% gsub(" ", "-", .)
  )

```

Scatter plot labeling candidate genes
```{r log2fc-scatter-candidate_genes, echo=FALSE, results='asis'}
label_genes <- use_data %>% filter(gene_name %in% candidate_genes)

res <- use_data %>% 
  ggplot(aes(log2FoldChange_ko, log2FoldChange_wt,
             color = -log10(padj_int), size = -log10(padj_int))) +
  geom_abline(lty = 'dashed', color = 'gray60', linewidth = one_pt/2) +
  geom_hline(yintercept = 0, lty = 'dashed', color = 'gray60', linewidth = one_pt/2) +
  geom_vline(xintercept = 0, lty = 'dashed', color = 'gray60', linewidth = one_pt/2) +
  geom_point(pch = 16, alpha = 0.7) +
  geom_text_repel(aes(label = gene_name), 
                  data = label_genes, 
                  min.segment.length = 0, 
                  seed = 42, 
                  box.padding = 0.5, 
                  max.overlaps = Inf, 
                  show.legend = FALSE, 
                  size = geom_text_size,
                  segment.size = one_pt/2,
                  segment.alpha = 0.7) +
  scale_color_viridis_c(direction = -1) +
  scale_size_continuous(range = c(0.1, 2)) +
  labs(
    title = paste('Differential expression analysis in', use_subset),
    y = 'log2FC shear stress in wt',
    x = 'log2FC shear stress in ko',
    color = '-log10(padj)\ninteraction',
    caption = paste('Adjusted P value capped at', p_limit)
  ) #+
  # scale_colour_gradientn(
  #   colours = c("#E6E9AF", "#8EA3A6", "#605EA1", "#22177A"),
  #   rescaler = ~ scales::rescale_mid(.x, mid = -log10(0.0001))
  # )

  subchunkify(
    print(res), 
    fig.width = 3.6, 
    fig.height = 3.6*0.618,
    dpi = 600,
    prefix = paste0('log2fc-scatter-candidate_genes-', file_prefix)
    # suffix = tolower(i) %>% gsub(" ", "-", .)
  )

```



```{r  log2fc-dotplot-candidate_genes, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

use_data <- foreach(i = use_names, .combine = rbind) %do% {
  dge_list[[i]] %>% 
    mutate(comparison = i)
}

# generate barplot for use_data using gene_names in candidate_genes, showing log2FoldChange and padj
comparison_palette <- c(chip_in_ko="#1F83B4",
                        chip_in_wt ="#FFBD4C",
                        ko_in_chip = "#29A03C",
                        ko_in_control = "#6F63BB",
                        interaction = "#E75727")
res <- use_data %>% 
  filter(gene_name %in% candidate_genes) %>%
  mutate(padj = ifelse(padj > p_limit, padj, p_limit)) %>%
  mutate(comparison = factor(comparison, levels = rev(names(comparison_palette)))) %>%
  ggplot(aes(y = comparison, x = log2FoldChange, 
             color = comparison, size = -log10(padj))) +
  geom_point(pch = 16) +
  geom_vline(xintercept = 0, lty = 2, linewidth =one_pt/2) +
  scale_size_continuous(range = c(0.05,2)) +
  facet_grid(rows = vars(gene_name)) +
  scale_color_manual(values = comparison_palette) +
  theme_facet +
  labs(x = 'log2 Fold Change', y = '')

  subchunkify(
    print(res), 
    fig.width = 3.6, 
    fig.height = 3.6*1.2,
    dpi = 600,
    prefix = paste0('log2fc-dotplot-candidate_genes', file_prefix)
    # suffix = tolower(i) %>% gsub(" ", "-", .)
  )
```
