---
title: "Shear stress CRISPR experiment"
subtitle: "Differential abundance using MAGeCK rra module for Brunello library"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  se_files: ./data/crispr/brunello_human_v1/se.rds
  output_dir: ./output/crispr/brunello_human_v1
  mageck_dir: mageck_rra
---


## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101

set.seed(use_seed)
```


Load packages
```{r load-libraries}
library(tidyverse)
library(foreach)
library(SummarizedExperiment)
library(DT)
library(colorblindr)
library(ggdendro)
library(RColorBrewer)
library(circlize)
library(Hmisc)
library(ComplexHeatmap)
library(ineq)
library(knitr)
library(kableExtra)
library(magrittr)
library(ggrepel)
library(ggpubr)
library(ggbeeswarm)
library(ggridges)
library(openxlsx)
library(MAGeCKFlute)
library(janitor)
```

Load ggplot theme
```{r ggplot-theme, include=TRUE}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/dge_report.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/gse_omnibus.r')
clean_msigdb_names <- function(x) x %>% gsub('HALLMARK_', '', .) %>% gsub('REACTOME_', '', .) %>% gsub('WP_', '', .) %>% gsub('BIOCARTA_', '', .) %>% gsub('KEGG_', '', .) %>% gsub('PID_', '', .) %>% gsub('GOBP_', '', .) %>% gsub('_', ' ', .)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

Load MSigDB gene sets
```{r load-msigdb}
msigdb_vs <- 'v2022.1.Hs'
gmt_files_symbols <- list(
  msigdb.c2.cp = paste0('./data/resources/MSigDB/', msigdb_vs, '/c2.cp.', msigdb_vs, '.symbols.gmt')
)
```

## Data procesing
Load Summarized Experiment object
```{r load-se}
se <- readRDS(params$se_file)
```

Combine all sublibraries and generate descriptive statistics
```{r combine-se}
se_chip <- se[!rowData(se)$control,se$condition == 'chip']
se_control <- se[!rowData(se)$control,se$condition == 'control']

mat_counts <- cbind(
  chip = rowSums(assay(se_chip, 'counts')),
  control = rowSums(assay(se_control, 'counts')) 
)

mat_counts_norm <- cbind(
  chip = rowSums(assay(se_chip, 'counts_norm_mageck')),
  control = rowSums(assay(se_control, 'counts_norm_mageck')) 
)


sec <- SummarizedExperiment(
  assays = list(counts = mat_counts, counts_norm = mat_counts_norm),
  colData = data.frame(sample_alias = 'chip', condition = 'chip') %>% 
    bind_rows(data.frame(sample_alias = 'control', condition = 'control')) %>% 
    DataFrame,
  rowData = rowData(se_chip)
)

rowData(sec) %<>% data.frame %>% 
  bind_cols(chip_sum = rowSums(assay(se_chip, 'counts_norm_mageck'))) %>% 
  bind_cols(control_sum = rowSums(assay(se_control, 'counts_norm_mageck'))) %>% 
  DataFrame
```

Generate addditional statistics for combined SE
```{r combine-stats}
sec$n_reads <-apply(assay(sec, 'counts'), 2, sum) 

sec$guides_over_1 <-apply(assay(sec, 'counts_norm') > 0, 2, sum) 
sec$guides_over_5 <-apply(assay(sec, 'counts_norm') >= 5, 2, sum) 
sec$guides_over_10 <-apply(assay(sec, 'counts_norm') >= 10, 2, sum)
sec$guides_over_30 <-apply(assay(sec, 'counts_norm') >= 30, 2, sum)
sec$guides_over_100 <-apply(assay(sec, 'counts_norm') >= 100, 2, sum)
sec$guides_over_500 <-apply(assay(sec, 'counts_norm') >= 500, 2, sum)

sec$p_guides_over_1 <-sec$guides_over_1 / nrow(sec)
sec$p_guides_over_5 <-sec$guides_over_5 / nrow(sec)
sec$p_guides_over_10 <-sec$guides_over_10 / nrow(sec)
sec$p_guides_over_30 <-sec$guides_over_30 / nrow(sec)
sec$p_guides_over_500 <-sec$guides_over_500 / nrow(sec)

# assay(sec, 'counts_norm') >= 10

# sec$gini_index <- assay(se, 'counts') %>% apply(., 2, ineq, type = 'Gini')
# sec$gini_index_log <- log10(assay(se, 'counts') + 1) %>% apply(., 2, ineq, type = 'Gini')

gene_table_stas <- rowData(sec) %>% data.frame %>% 
  group_by(gene) %>% 
  summarise(
    n_chip_over_1 = sum(chip_sum > 1),
    n_chip_over_5 = sum(chip_sum > 5),
    n_chip_over_10 = sum(chip_sum > 10),
    n_chip_over_30 = sum(chip_sum > 30),
    n_chip_over_100 = sum(chip_sum > 100),
    n_chip_over_500 = sum(chip_sum > 500),
    n_control_over_1 = sum(control_sum > 1),
    n_control_over_5 = sum(control_sum > 5),
    n_control_over_10 = sum(control_sum > 10),
    n_control_over_30 = sum(control_sum > 30),
    n_control_over_100 = sum(control_sum > 100),
    n_control_over_500 = sum(control_sum > 500)
  ) 
  
gene_table <- rowData(sec) %>% data.frame %>% 
  group_by(gene) %>% 
  summarise(
    n_chip_over_1 = sum(chip_sum > 1),
    n_chip_over_5 = sum(chip_sum > 5),
    n_chip_over_10 = sum(chip_sum > 10),
    n_chip_over_30 = sum(chip_sum > 30),
    n_chip_over_100 = sum(chip_sum > 100),
    n_chip_over_500 = sum(chip_sum > 500),
    n_control_over_1 = sum(control_sum > 1),
    n_control_over_5 = sum(control_sum > 5),
    n_control_over_10 = sum(control_sum > 10),
    n_control_over_30 = sum(control_sum > 30),
    n_control_over_100 = sum(control_sum > 100),
    n_control_over_500 = sum(control_sum > 500)
  ) %>% 
  mutate(
    depleted_chip = n_chip_over_10 < 2,
    depleted_control = n_control_over_10 < 2
  ) %>% 
  dplyr::select(gene, starts_with('depleted'), ends_with('over_10'), everything())

```



## Configure analyses
List of comparisons
```{r conf-comparisons}
x <- colData(se) %>% data.frame

comp_list <- list(
  sublibrary_1 = list(
    x %>% filter(sg_rna_sublibrary == 1 & condition == 'chip') %>% pull(sample_alias),
    x %>% filter(sg_rna_sublibrary == 1 & condition == 'control') %>% pull(sample_alias),
    rowData(se) %>% data.frame %>% filter(sublibrary %in% c("1", "control") ) %>% rownames
    ),
  sublibrary_2 = list(
    x %>% filter(sg_rna_sublibrary == 2 & condition == 'chip') %>% pull(sample_alias),
    x %>% filter(sg_rna_sublibrary == 2 & condition == 'control') %>% pull(sample_alias),
    rowData(se) %>% data.frame %>% filter(sublibrary %in% c("2", "control") ) %>% rownames
    ),
  sublibrary_3 = list(
    x %>% filter(sg_rna_sublibrary == 3 & condition == 'chip') %>% pull(sample_alias),
    x %>% filter(sg_rna_sublibrary == 3 & condition == 'control') %>% pull(sample_alias),
    rowData(se) %>% data.frame %>% filter(sublibrary %in% c("3", "control") ) %>% rownames
    ),
  sublibrary_4 = list(
    x %>% filter(sg_rna_sublibrary == 4 & condition == 'chip') %>% pull(sample_alias),
    x %>% filter(sg_rna_sublibrary == 4 & condition == 'control') %>% pull(sample_alias),
    rowData(se) %>% data.frame %>% filter(sublibrary %in% c("4", "control") ) %>% rownames
    ),
  sublibrary_5 = list(
    x %>% filter(sg_rna_sublibrary == 5 & condition == 'chip') %>% pull(sample_alias),
    x %>% filter(sg_rna_sublibrary == 5 & condition == 'control') %>% pull(sample_alias),
    rowData(se) %>% data.frame %>% filter(sublibrary %in% c("5", "control") ) %>% rownames
    )
)
```

```{r conf-comparisons-print}
data_comp <- foreach(i = names(comp_list), .combine = rbind) %do% {
  c(
    Comparison = i,
    Case = paste(comp_list[[i]][[1]], collapse = " "),
    Control = paste(comp_list[[i]][[2]], collapse = " ")
  )
}

data_comp %>% 
  datatable(., 
            rownames = FALSE, 
            filter = 'top', 
            caption = 'Comparison of guide abundance. P value by two-sided Wiloxon test',
            extensions = 'Buttons', 
            options = list(
              dom = 'Blfrtip',
              buttons = c('csv', 'excel'),
              columnDefs = list(
                list(width = 120, targets = 1:2)
              )
            )
            )

```



## Run MAGeCK

Generate MAGeCK test scripts. Run MAGeCK test for each comparison in the terminal
```{r generate-mageck-cmds, eval = FALSE}
out_dir <- file.path(params$output_dir, params$mageck_dir)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# Generate list of control sgRNA IDs
non_target_list <- rowData(se) %>% data.frame %>% filter(control) %>% rownames %>% data.frame
write_tsv(non_target_list, file = file.path(out_dir, 'control_sgrna.txt'), col_names = F)
  
# Generate files by comparison
res_cmd <- foreach(i = names(comp_list), .combine = rbind) %do% {
  use_comp <- comp_list[[i]]
  
  # Generate count matrix
  count_mat_name <- paste0(i,'.counts.txt')
  count_mat <- assay(se[use_comp[[3]],], 'counts') %>% 
    data.frame(check.names = FALSE) %>% 
    rownames_to_column('sgrna_name') %>% 
    left_join(rowData(se) %>% data.frame %>% select(sgrna_name, gene)) %>% 
    dplyr::select(sgrna_name, gene, everything())
  
  write_tsv(count_mat, file = file.path(out_dir, count_mat_name))
  
  # Generate mageck commands
   paste('mageck test -k', count_mat_name, '--control-sgrna control_sgrna.txt --norm-method control -t', paste(comp_list[[i]][[1]], collapse = ','),'-c',  paste(comp_list[[i]][[2]], collapse = ','), '-n', i, '\n')
} %>% data.frame()

res_cmd <- rbind('conda activate mageckenv', res_cmd, 'conda deactivate')
write_tsv(res_cmd, file.path(out_dir, 'run_mageck_test.sh'), col_names = FALSE)
```

Load MAGeCK results
```{r load-mageck}
gene_summ_files <- list.files(path = file.path(params$output_dir, params$mageck_dir), pattern = 'gene_summary.txt', full.names = TRUE)
analysis_prefix <- basename(gene_summ_files) %>% gsub(".gene_summary.txt", "", .)
gene_summ <- foreach(i = gene_summ_files) %do% {read.delim(i, check.names = FALSE) %>% clean_names()}
names(gene_summ) <- analysis_prefix

sgrna_summ_files <- list.files(path = file.path(params$output_dir, params$mageck_dir), pattern = 'sgrna_summary.txt', full.names = TRUE)
analysis_prefix <- basename(sgrna_summ_files) %>% gsub(".sgrna_summary.txt", "", .)
sgrna_summ <- foreach(i = sgrna_summ_files) %do% {read.delim(i, check.names = FALSE) %>% clean_names()}
names(sgrna_summ) <- analysis_prefix
```

Combine sublibrary results
```{r combine-mageck}
gene_summ$combined <- foreach(i = names(gene_summ), .combine = rbind) %do% {
  gene_summ[[i]] %>% 
    mutate(sublibrary = i)
}

sgrna_summ$combined <- foreach(i = names(sgrna_summ), .combine = rbind) %do% {
  sgrna_summ[[i]] %>% 
    mutate(sublibrary = i)
}


# recalculate FDR after combining sublibraries
gene_summ$combined  %<>% 
  mutate(
    neg_fdr = p.adjust(neg_p_value, 'BH'),
    pos_fdr = p.adjust(pos_p_value, 'BH')
  )

sgrna_summ$combined %<>% 
  mutate(
    fdr = p.adjust(p_twosided, 'BH')
  )


# recalculate rank
gene_summ$combined  %<>% 
  mutate(
    neg_rank = rank(neg_p_value, ties.method = 'first'),
    pos_rank = rank(pos_p_value, ties.method = 'first')
  )

```

## Run GSE omnibus
```{r run-gse-omnibus, eval = FALSE}
library(org.Hs.eg.db)
x <- gene_summ$combined
gse_res <- gse_omnibus(
    feature_names = x$id,
    p = x$neg_fdr,
    fc = x$neg_lfc,
    fc_thrs = 0.1,
    gmt_files = gmt_files_symbols,
    
    run_enricher = TRUE,

    args_enricher = list(minGSSize = 10, maxGSSize = 2500, pvalueCutoff = 1), 
    )
saveRDS(gse_res, file = file.path(params$output_dir, params$mageck_dir, 'gse.rds'))
```

```{r load-gse-omnibus}
gse_res <- readRDS(file.path(params$output_dir, params$mageck_dir, 'gse.rds'))
```




## Extended Data Figure 5a: CRISPR screen data quality control

Bar plots showing the percentage of detected gRNAs in samples incubated in static conditions (control) or circulated through the microfluidic platform (shear stress; left). Density plots showing
the distribution of read counts of gRNAs retrieved in control (n = 1) and shear stress (n = 1) conditions. Dashed line indicates the detection threshold (10 counts) (right).

```{r combine-stats-guides-barplot, fig.width = 1.85,  fig.asp = 0.76}
use_linewidth <- 1/2.141959
colData(sec) %>% data.frame %>%
  mutate(
    sample_alias = str_to_title(sample_alias),
    sample_alias = factor(sample_alias, levels = c('Control', 'Chip'))
    ) %>% 
  ggplot(aes(sample_alias, p_guides_over_10)) +
  geom_col() +
  geom_text(aes(label = round(100*p_guides_over_10, 1)), 
            vjust = 1.5, 
            colour = "white", 
            size = geom_text_size) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01)),
                     labels = scales::percent) +
  labs(
    x = '',
    y = 'Percentage of guides detected'
  ) +
  theme(
    axis.ticks = element_line(linewidth = use_linewidth/2),
    axis.ticks.x = element_blank(),
    axis.line = element_line(linewidth = use_linewidth/2)
  )

```

```{r combine-stats-grna-dist, fig.width = 1.85, fig.asp = 0.500}
use_data <- cbind(
  rowData(sec) %>% data.frame %>% dplyr::select(sublibrary),
  assay(sec, 'counts')
) %>% 
  pivot_longer(
    cols = all_of(c('chip', 'control')),
    names_to = 'condition', 
    values_to = 'counts') %>% 
  mutate(
    condition = ifelse(condition == 'control', 'Control', 'Shear stress')
  )

use_linewidth <- 1/2.141959
use_data %>% 
  ggplot(aes(counts + 1, fill = condition)) +
    geom_density(alpha = 0.8, linewidth =  use_linewidth/2) +
    geom_vline(xintercept = 11, lty = 3, linewidth =  use_linewidth/2) +
    scale_fill_manual(values = c(Control = 'grey80', `Shear stress` = '#D55E00')) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.03))) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.01)), trans = 'log10') +
    labs(
      x = 'Number of reads + 1',
      fill = ''
    )+
    theme(
      axis.ticks = element_line(linewidth = use_linewidth/2),
      axis.line = element_line(linewidth = use_linewidth/2),
      axis.text = element_text(size = 4),
      axis.title = element_text(size = 4),
      legend.text = element_text(size = 4),
      legend.key.width = unit(0.1, "cm")
    )
```

## Extended Data Figure 5b: CRISPR screen data quality control
Over-Representation analysis (ORA) plot of genes depleted in static conditions. FDR, false discovery rate.
```{r ora, eval = FALSE}
x <- gene_table %>% 
  mutate(
    detected = n_chip_over_10 >= 2,
    p = ifelse(!detected, 0.01, 1),
    fc = ifelse(!detected,2, 0)
    )

ora_fitness <- gse_omnibus(
    feature_names = x$gene,
    p = x$p,
    fc = x$fc,
    fc_thrs = 0.1,
    gmt_files = gmt_files_symbols,
    
    run_enricher = TRUE,
    run_enrichGO = TRUE,

    return_sets = 'abs',
    go_ontologies = c('BP'),
    go_simplify = TRUE,
    go_simplify_p = 0.05,
    go_simplify_cutoff = 0.4,
    args_enricher = list(minGSSize = 10, maxGSSize = 250, pvalueCutoff = 1), 
    annot_db = 'org.Hs.eg.db',
    organism = 'hsapiens'
    )
saveRDS(ora_fitness, file = file.path(params$output_dir, 'ora_fitness.rds'))
```

GO-BP, gene sets with an adjusted P < 0.05
```{r ora-go, fig.width=2, fig.asp = 0.9}
ora_fitness <- readRDS(file.path(params$output_dir, 'ora_fitness.rds'))
use_linewidth <- 1/2.141959
ora_fitness$enrichGO$abs$BP@result %>% 
  filter(simplify) %>% 
  filter(p.adjust < 0.05) %>% 
  mutate(Description = fct_reorder(Description, -log10(p.adjust))) %>%
  ggplot(aes(-log10(p.adjust), Description, 
             color = -log10(p.adjust),
             size = GeneProp)) +
  geom_point(shape = 19) +
  scale_color_gradient(
     high = "#132B43",
     low = "#56B1F7"
  ) +
  scale_y_discrete(labels = default_labeller(50)) +
  scale_x_continuous(expand = expansion(mult = c(0.1, 0.1))) +
  scale_size_continuous(range=c(0.8, 2)) +
  labs(y = '', 
       x = '',  
       size = expression(paste("lo", g[10],"(FDR)")),
       color = 'FDR < 0.05'
       ) +
  theme(
    text = element_text(size = 5),
    axis.line = element_blank(),
    axis.ticks.y = element_blank(),
    # axis.ticks.y = element_line(linewidth = use_linewidth/2),
    axis.ticks.x = element_line(linewidth = use_linewidth/2),
    panel.border = element_rect(color = "black", fill = NA, linewidth = use_linewidth),
   
    strip.background = element_blank(),
    axis.text.y = element_text(size = 3),
    axis.text.x = element_text(size = 3),
    legend.text = element_text(size = 2),
    legend.title = element_text(size = 2),
    legend.key.width = unit(0.1, "cm")
  )
```



## Figure 3b: Rank plot

Plot presenting genes ranked by log2 fold-change with top depleted genes under shear stress highlighted in blue. All other genes are indicated in grey. Thresholds (dashed lines) were calculated using the mean ± 5 standard deviations of all log2 fold-changes treatment scores. 

```{r rank-plot-raw, fig.width = 1.35, fig.asp = 1.1}
use_linewidth <- 1/2.141959
i <- 'combined'
x <- gene_summ[[i]] %>%  ReadRRA(score = 'lfc') %>% dplyr::rename(LFC = Score)
gene_list <-  x$LFC %>% set_names( x$id) %>% na.omit
# RankView(gene_list, top = 20, bottom = 0)

# RankView(
#   gene_list,
#   top = 0,
#   bottom = 0)


RankView2 <- function (rankdata, genelist = NULL, decreasing = TRUE, top = 5, 
    bottom = 5, cutoff = 2, main = NULL, filename = NULL, width = 5, 
    height = 4, label.size = 2.5, ...) 
{
    requireNamespace("ggrepel", quietly = TRUE) || stop("need ggrepel package")
    if (length(cutoff) < 2) {
        cutoff = CutoffCalling(rankdata, cutoff)
        cutoff = c(-cutoff, cutoff)
    }
    else cutoff = sort(cutoff)
    data = data.frame(Gene = names(rankdata), diff = rankdata, 
        stringsAsFactors = FALSE)
    if (decreasing) 
        data$Rank = rank(-data$diff)
    else data$Rank = rank(data$diff)
    data$group = "no"
    data$group[data$diff > cutoff[2]] = "up"
    data$group[data$diff < cutoff[1]] = "down"
    return(data)
}

data_rank<- RankView2(
  gene_list,
  cutoff = 5
)

mycolour <-  c(no = "gray80", up = "#e41a1c", down = "#377eb8")

ggplot(data_rank, aes(x = Rank, y = diff, color = group)) +
  geom_point(size = 0.2) +
  geom_hline(
    yintercept =  CutoffCalling(gene_list, 5),
    lty = 2,
    linewidth = use_linewidth/2
    ) +
  geom_hline(
    yintercept =  -CutoffCalling(gene_list, 5),
    lty = 2,
    linewidth = use_linewidth/2
  ) +
  geom_hline(
    yintercept =  0,
    lty = 1,
    linewidth = use_linewidth/2
  ) +
  scale_color_manual(values = mycolour) +
  labs(x = "Rank", y = expression(paste("lo", g[2],"(Fold change)"))) +
  theme(
    legend.position="none",
    text = element_text(size = 4),
    axis.text = element_text(size = 4),
    axis.line = element_line(linewidth = use_linewidth/2),
    axis.ticks = element_line(linewidth = use_linewidth/2)
  )
```


## Figure 3c: Volcano plot

Volcano plot illustrating the results of the CRISPR screen. Genes significantly depleted in samples circulated through the microfluidic platform (shear stress) are indicated in blue (FDR < 0.05). Non-significant genes are depicted in grey.

```{r volcano-plot, fig.width = 1.35,  fig.asp = 0.86}
i <- 'combined'
x <- gene_summ[[i]] %>%  ReadRRA(score = 'lfc') %>% dplyr::rename(LFC = Score)
use_linewidth <- 1/2.141959
topn <- 20

# Set coloring group
mycolour <-  c(no = "gray80", up = "#e41a1c", down = "#377eb8")
x %<>% 
  mutate(
    group = ifelse(LFC < 0 & FDR < 0.05, 'down', 'no'),
    group = ifelse(LFC > 0 & FDR < 0.05, 'up', group)
  )

# Symmetric x-limits
use_xlim <- c(-max(abs(x$LFC), na.rm = TRUE), max(abs(x$LFC), na.rm = TRUE)) %>% 
  round

x %>% 
  ggplot(aes(LFC, -log10(FDR), color = group)) +
  geom_point(size = 0.4, alpha = 0.6, shape = 16) +
  scale_color_manual(values = mycolour) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 4)) +
  scale_x_continuous(expand = c(0, 0),
                     limits = use_xlim) +
  labs(
    x = expression(paste("lo", g[2],"(Fold change)")),
    y = expression(paste("-lo", g[10],"(FDR)"))
  ) +
  theme(
    legend.position="none",
    text = element_text(size = 4),
    axis.text = element_text(size = 4),
    axis.line = element_line(linewidth = use_linewidth/2),
    axis.ticks = element_line(linewidth = use_linewidth/2)
  )

```


## Figure 3d and Extended Data Figure 5c,d: Functional enrichment

```{r ora-msigdb-c2_cp-dot-conf, fig.width=2, fig.asp = 0.9}
library(enrichplot)
top_n <- 20
n_split <- 8
p_thresh <- 0.00001
use_gse <- gse_res$enricher$down$msigdb.c2.cp
use_gse <- pairwise_termsim(use_gse)

# Keep only reactome and kegg and WP, removing overlapping pathways, selecting P.adjust
res_top <- use_gse@result %>% 
  filter(grepl('REACTOME|KEGG|WP_', ID)) %>% 
  filter(setSize < 500) %>% 
  filter(!ID %in% 'KEGG_CELL_CYCLE') %>%
  filter(!ID %in% 'WP_DNA_REPLICATION') %>%
  filter(!ID %in% 'KEGG_PYRIMIDINE_METABOLISM') %>%
  mutate(
    Description = clean_msigdb_names(Description)
  ) %>%
  filter(p.adjust < p_thresh)

use_gse@result <- use_gse@result[rownames(res_top),]

# Similarity matrix
similarity_matrix <- use_gse@termsim
use_mat <- similarity_matrix[rownames(res_top), rownames(res_top)]
for(x in rownames(use_mat)){
  for(y in colnames(use_mat)) {
    if(x == y) {
       use_mat[x,y] <- 1
    } else {
      max_sim <- max(c(use_mat[x,y], use_mat[y,x]), na.rm = TRUE)
      use_mat[x,y] <- max_sim
      use_mat[y,x] <- max_sim
    }
  }
}
rownames(use_mat) <- clean_msigdb_names(rownames(use_mat))
colnames(use_mat) <- clean_msigdb_names(colnames(use_mat))

# Cluster the similarity matrix
# Use the ward.D method to avoid overlapping ancestor nodes of each group
hc <- hclust(
  as.dist(1- use_mat),
  method = 'ward.D'
)

split <- cutree(hc, k=n_split) %>% data.frame %>% set_names('cluster')
split$curated_id <- clean_msigdb_names(rownames(split))

# Combine cell cycle gene sets into one cluster
rho_cluster <- split %>% filter(curated_id == 'RHO GTPASE EFFECTORS') %>% pull(cluster)
split %<>%
  mutate(
    cluster = ifelse(
      cluster == rho_cluster &
      curated_id %in% c('MITOTIC METAPHASE AND ANAPHASE',
                        'MITOTIC SPINDLE CHECKPOINT',
                        'CELL CYCLE CHECKPOINTS',
                        'SEPARATION OF SISTER CHROMATIDS'
                        ),
      2,
      cluster
    )
  )
# Manually annotate clusters, using top gene sets per cluster as guide
# fake_gseaResult@result %>% filter(cluster == '1') %>% select(ID, NES, p.adjust, GeneRatio, core_enrichment)
split$cluster_name <- case_when(
  split$cluster == 1 ~ 'mRNA splicing',
  split$cluster == 2 ~ 'Cell cycle - M phase',
  split$cluster == 3 ~ 'Rho GTPases - CDC42',
  split$cluster == 4 ~ 'Proteolisis and RNA degradation',
  split$cluster == 5 ~ 'Transport of mature mRNA',
  split$cluster == 6 ~ 'DNA replication',
  split$cluster == 7 ~ 'SUMO dependent protein transport',
  split$cluster == 8 ~ 'rRNA processing',
  TRUE ~ 'Other'
)


split$cluster_arranged <- case_when(
  split$cluster_name == 'mRNA splicing' ~ 3,
  split$cluster_name == 'Cell cycle - M phase' ~ 2,
  split$cluster_name == 'Rho GTPases - CDC42' ~ 1,
  split$cluster_name == 'Proteolisis and RNA degradation' ~ 5,
  split$cluster_name == 'Transport of mature mRNA' ~ 7,
  split$cluster_name == 'DNA replication' ~ 4,
  split$cluster_name == 'SUMO dependent protein transport' ~ 6,
  split$cluster_name == 'rRNA processing' ~ 8,
  TRUE ~ NA
)
split %<>% 
  mutate(
    cluster_fct = factor(cluster, levels = as.character(1:n_split)),
    cluster_name =  fct_reorder(cluster_name, cluster_arranged)
  )


res_top <- res_top %>% 
  left_join(split, by = c('Description' = 'curated_id'))
```


### Figure 3d and Extended Data Figure 5d: Similarity matrix

Similarity matrix illustrating significantly enriched pathways within the depleted genes in samples circulated through microfluidic platform (shear stress) compared to samples incubated in static conditions (control), calculated based on the Jaccard index to measure pathway semantic overlap including the extensive list of pathways.

```{r ora-msigdb-c2cp-similarity-matrix-heatmap-wardD, fig.width=7.2}
col_fun <- colorRamp2(seq(0, 1, length.out = 4), brewer.pal(4, "GnBu"))
ht <- Heatmap(
  use_mat, 
  name = 'Jaccard index',
  # column_split = n_split,
  # row_split = n_split,
  # column_km = n_split,
  # row_km = n_split,
  column_split = split$cluster_arranged,
  row_split = split$cluster_name,
  cluster_column_slices = FALSE,
  cluster_row_slices = FALSE,
  
  # column_title = NULL,
  # row_title = NULL,
  row_title_rot = 0,
  row_title_gp = gpar(fontsize = 5),
  column_title_gp = gpar(fontsize = 5),
  
  col = col_fun,
  show_column_dend = FALSE,
  show_column_names = FALSE, 
  show_row_dend = FALSE,
  row_names_gp = gpar(fontsize = 3),
  border = TRUE,
  # top_annotation = ha_top, 
  heatmap_legend_param = list(
    title_gp = gpar(fontsize = 8),
    labels_gp = gpar(fontsize = 8),
    direction = "horizontal"),
  width = unit(7, "cm")
  )
draw(ht,  heatmap_legend_side = "bottom")
```

### Extended Data Figure 5c

Over Representation Analysis (ORA) plot for gene sets with an adjusted P value < 0.05.

```{r ora-msigdb-c2cp-dotplot-full, fig.width = 3, fig.asp = 2.0}
use_linewidth <- 1/2.141959

use_data <- res_top %>% 
  mutate(
    p.sig = p.adjust < 0.05,
    p.adjust = -log10( p.adjust),
    Description = fct_reorder(Description, p.adjust)
    )
use_data %>% 
  ggplot(aes(p.adjust, Description, color = p.adjust, size = GeneProp)) +
  geom_point(shape = 16) +
  facet_grid(rows = vars(cluster_name),
             scale = 'free', space = 'free',
             labeller = label_wrap_gen(40) ) +
  scale_color_distiller(palette = "BuPu", na.value = "grey70", 
                        direction = 1,
                        limits = c(0, max(use_data$p.adjust)))+
  # scale_color_manual(values = c('white', 'black')) +
  scale_y_discrete(labels = default_labeller(50)) +
  scale_size_continuous(range=c(0.1, 0.9)) +
  labs(y = '', 
       x = '',  
       size = expression(paste("lo", g[10],"(FDR)")),
       ) +
  theme(
    text = element_text(size = 4),
    axis.line = element_blank(),
    # axis.ticks.y = element_blank(),
    axis.ticks.y = element_line(linewidth = use_linewidth/2),
    axis.ticks.x = element_line(linewidth = use_linewidth/2),
    panel.border = element_rect(color = "black", fill = NA, linewidth = use_linewidth),
   
    strip.background = element_blank(),
    axis.text.y = element_text(size = 2),
    axis.text.x = element_text(size = 3),
    legend.text = element_text(size = 2),
    legend.title = element_text(size = 2),
    legend.key.width = unit(0.1, "cm"),
    
    strip.text.y = element_text(size = 2, angle = 0, hjust = 0),
    panel.spacing = unit(0.07, "lines")
  )

```


## Figure 3e: CDC42 sgRNA abundance

Normalized *CDC42* gRNA abundance comparison in samples incubated in static conditions (control) or circulated through the microfluidic platform for four hours (shear stress). Each dot corresponds to an individual gRNA. The lines connect the paired values.

```{r guides-paired-abundance}
use_data <- sgrna_summ$combined %>% 
  dplyr::select(sgrna, gene, control_count, treatment_count) %>% 
  pivot_longer(-c(sgrna, gene), names_to = 'condition', values_to = 'counts') %>% 
  mutate(
    condition = ifelse(condition == 'control_count', 'Control', 'Chip'),
    condition = factor(condition, levels = c('Control', 'Chip'))
  )
```

```{r guides-paired-abundance-cdc42, fig.width = 1.8, fig.asp = 1.2}
use_gene <- 'CDC42'
use_data %>% 
  filter(gene == use_gene) %>% 
  ggplot(aes(x = condition, y = counts, group = sgrna)) +
  geom_point() +
  geom_line() +
  scale_x_discrete(expand = c(0.08, 0.08)) +
  scale_y_continuous(expand = c(0.049, 0)) +
  labs(
    x = '',
    y = 'Normalized counts',
    title = paste('sgRNAs in', use_gene)
  )
```
