---
title: "Analysis of all models"
subtitle: "Combined figures"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  output_dir: ./output/rnaseq/bulk/differential_expression/all_models
  k_predefined: '8'
  ncpu: 4
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
# library(tidyverse)
library(ggplot2)
library(tibble)
library(tidyr)
library(readr)
library(purrr)
library(dplyr)
library(forcats)
library(magrittr)
library(foreach)

library(showtext)
library(ggupset)
library(ggtext)
library(glue)
library(cowplot)
library(colorblindr)

library(DT)
library(knitr)
library(kableExtra)

library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(ComplexHeatmap)
library(circlize)
library(openxlsx)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(scuttle)
library(scran)
library(pheatmap)
library(PCAtools)
library(apeglm)
library(patchwork)
library(geomtextpath)
library(enrichplot)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/dge_report.r')
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/gse_treeplot.R')
source('./code/R-functions/wordcloud.R') # required for gse_treeplot.R
clean_msigdb_names <- function(x) x %>% gsub('HALLMARK_', '', .) %>% gsub('REACTOME_', '', .) %>% gsub('WP_', '', .) %>% gsub('BIOCARTA_', '', .) %>% gsub('KEGG_', '', .) %>% gsub('PID_', '', .) %>% gsub('GOBP_', '', .) %>% gsub('_', ' ', .)
```

Load MSigDB gene sets
```{r load-msigdb}
msigdb_vs <- 'v2022.1.Hs'
gmt_files_symbols <- list(
  msigdb.c2.cp = paste0('./data/resources/MSigDB/', msigdb_vs, '/c2.cp.', msigdb_vs, '.symbols.gmt')
)
```

## Data processing

### Load original data
```{r load-se-data, eval = FALSE}
se_br16 <- readRDS('data/rnaseq/bulk/se_br16.rds')
se_brx07 <- readRDS('data/rnaseq/bulk/se_brx07.rds')
se_brx50 <- readRDS('data/rnaseq/bulk/se_brx50.rds')

se_br16$sample_alias <- paste0(se_br16$donor, '_',se_br16$sample_alias)
se_brx07$sample_alias <- paste0(se_brx07$donor, '_',se_brx07$sample_alias)
se_brx50$sample_alias <- paste0(se_brx50$donor, '_',se_brx50$sample_alias)

colnames(se_br16) <- se_br16$sample_alias
colnames(se_brx07) <- se_brx07$sample_alias
colnames(se_brx50) <- se_brx50$sample_alias

rowData(se_br16) %<>% data.frame %>% dplyr::select(-mean, -detected) %>% DataFrame
rowData(se_brx07) %<>% data.frame %>% dplyr::select(-mean, -detected) %>% DataFrame
rowData(se_brx50) %<>% data.frame %>% dplyr::select(-mean, -detected) %>% DataFrame

se <- cbind(se_br16, se_brx07, se_brx50)
se_o <- colData(se) %>% data.frame %>% 
  mutate(
    total_features = nrow(se),
    pct_detected = 100*detected/total_features
  )

rm(se_br16, se_brx07, se_brx50)
```

### Data wrangling

Modify annotation
```{r dw-coldata, eval = FALSE}
colData(se) <- colData(se) %>% data.frame %>% 
  mutate(
    condition = factor(condition, levels= c('Control', 'Chip')),
    time = factor(time, levels=time %>% as.numeric %>% sort %>% unique),
    group = paste(condition, time, sep = '_'),
    group = factor(group, 
                   levels = paste(
                     c('Control', 'Chip'), 
                     time %>% as.character %>% as.numeric %>% sort %>% unique %>% rep(2) %>% sort, 
                     sep = '_')
                   )
    
  ) %>% 
  DataFrame
```

Subset by feature type. Using protein coding features and lncRNA 
```{r dw-subset-feature-type, eval = FALSE}
use_rows <- rowData(se)$gene_type %in% c('protein_coding', 'lncRNA')
se <- se[use_rows,]
```

Filtering out low-abundance features. Keeping those features with at least 10 counts in at least 70% of samples of the smallest group.
```{r dw-subset-feature-abundance, eval = FALSE}
use_rows <- filterByExpr(se, group=paste(se$condition,se$time, sep = '-'), min.count = 10, min.total.count = 15, min.prop = 0.7)
se <- se[use_rows,]

# Update PerCellQCMetrics
colData(se) <- colData(se) %>% data.frame %>% dplyr::select(-(sum:subsets_Ribo_percent)) %>% DataFrame
se <- addPerCellQCMetrics(
  se,
  subsets=list(
    Mito=which(rowData(se)$is.mito),
    Ribo=which(rowData(se)$is.ribo)
  ),
  percent.top = c(1,5,10,20, 50, 100),
  threshold = 3
  )

# Table of removed features
table(use_rows) %>% 
  data.frame %>% 
  mutate(use_rows = ifelse(use_rows == TRUE, 'keep', 'remove')) %>% 
  column_to_rownames('use_rows') %>% 
  kbl(caption = 'Filtered features') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

TMM normalization
```{r dw-tmm, eval = FALSE}
norm_method <- "TMM"
exprs_mat_for_norm <- assay(se, 'counts')
norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
lib_size <- colSums(exprs_mat_for_norm)/10^6
se$size_factor <- (norm_factors * lib_size)
se <- logNormCounts(se, size.factors =se$size_factor)
se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)
```

CPM calculation
```{r dw-cpm, eval = FALSE}
assay(se, 'cpm') <- edgeR::cpm(assay(se, 'normcounts'))
assay(se, 'logcpm') <- edgeR::cpm(assay(se, 'normcounts'), log = TRUE)
```

Save final data
```{r dw-save-se, eval = FALSE}
saveRDS(se, file = file.path(params$output_dir, 'se_merged.rds'))
```

Load final data
```{r dw-loa-se}
se <- readRDS(file = file.path(params$output_dir, 'se_merged.rds'))
```



### Differential expression
Here we load differential expression results for each model. This results have been pre-computed in the corresponding Rmd files:

- analysis/rnaseq-bulk-br16-time_series.Rmd
- analysis/rnaseq-bulk-brx07-time_series.Rmd
- analysis/rnaseq-bulk-brx50-time_series.Rmd

```{r dge-donor-conf, eval = FALSE}
dir_output <- dirname(params$output_dir)
dge <- foreach(i=unique(se$donor)) %do% readRDS(file = file.path(dir_output, tolower(i), 'dge_deseq2.rds'))
names(dge) <- unique(se$donor)

# fold-change data.frame
fc_by_time <- foreach(tm = levels(se$time)[-1]) %do% {
  foreach(donor = names(dge), .combine = rbind) %do% {
    dge[[donor]]$lrt_time[[tm]] %>% dplyr::select(ensembl_gene_id, gene_name, log2FoldChange) %>% mutate(donor = donor)
  } %>% data.frame(check.names = FALSE) %>% 
    pivot_wider(names_from = donor, values_from = log2FoldChange)
}
names(fc_by_time) <- levels(se$time)[-1]

# padj data.frame
p_by_time <- foreach(tm = levels(se$time)[-1]) %do% {
  foreach(donor = names(dge), .combine = rbind) %do% {
    dge[[donor]]$lrt_time[[tm]] %>% dplyr::select(ensembl_gene_id, gene_name, padj) %>% mutate(donor = donor)
  } %>% data.frame(check.names = FALSE)
}
names(p_by_time) <- levels(se$time)[-1]

# Single data.frame with all the DGE results (all models and times)
dge_df <- foreach(tm = levels(se$time)[-1], .combine = rbind) %do% {
  foreach(donor = names(dge), .combine = rbind) %do% {
    dge[[donor]]$lrt_time[[tm]] %>% mutate(donor = donor)
  } %>% data.frame(check.names = FALSE) %>% 
    mutate(time = tm)
}

saveRDS(dge_df, file = file.path(params$output_dir, 'dge_deseq2_df_combined.rds'))
```

Define list and expression matrix used for heatmap and module definition
```{r dge-donor-topgenes, eval = FALSE}
# Logcounts matrix of top-genes
use_rows <- dge_df %>% 
  filter(padj < 0.01) %>% 
  filter(ensembl_gene_id %in% rownames(se)) %>% 
  pull(ensembl_gene_id) %>% 
  unique
mat <- assay(se[use_rows,], 'logcounts')


zmat <- foreach(donor = unique(se$donor), .combine = cbind) %do% {
  use_se <- se[,se$donor == donor]
  use_mat <- mat[,colnames(use_se)]
  res <- t(apply(use_mat, 1, function(x) {scale(x, center = mean(x[colnames(use_se)[use_se$group == 'Control_0']]), scale = TRUE)}))
  res[is.na(res)] <- 0
  colnames(res) <- colnames(use_mat)
  res
}


# Order columns by condition - time - replicate
col_ord <- colData(se) %>% data.frame %>% arrange(condition, time, replicate) %>% pull(sample_alias)
mat <- mat[,col_ord]
zmat <- zmat[,col_ord]

# Save matrix
mat_topgenes <- mat
zmat_topgenes <- zmat

rm(mat, zmat)
saveRDS(mat_topgenes, file = file.path(params$output_dir, 'dge_deseq2_combined_mat_topgenes.rds'))
saveRDS(zmat_topgenes, file = file.path(params$output_dir, 'dge_deseq2_combined_zmat_topgenes.rds'))
```

Load data
```{r dge-deseq2-lrt-load}
dge_df <- readRDS(file = file.path(params$output_dir, 'dge_deseq2_df_combined.rds'))
mat_topgenes <- readRDS(file.path(params$output_dir, 'dge_deseq2_combined_mat_topgenes.rds'))
zmat_topgenes <- readRDS(file.path(params$output_dir, 'dge_deseq2_combined_zmat_topgenes.rds'))
```

### Functional enrichment

```{r gse-omnibus-time-compile, eval = FALSE}
dir_output <- dirname(params$output_dir)
time_gse_omnibus <- foreach(i=unique(se$donor)) %do% readRDS(file = file.path(dir_output, tolower(i), 'dge_deseq2_time_gse.rds'))
names(time_gse_omnibus) <- unique(se$donor)

saveRDS(time_gse_omnibus, file = file.path(params$output_dir, 'time_gse_omnibus.rds'))
```

```{r gse-omnibus-time-load}
time_gse_omnibus <- readRDS(file.path(params$output_dir, 'time_gse_omnibus.rds'))
```


## Supplementary tables
```{r st1-load-data}
dir_output <- dirname(params$output_dir)
dge_list <- foreach(i=unique(se$donor)) %do% readRDS(file = file.path(dir_output, tolower(i), 'dge_deseq2.rds'))
names(dge_list) <- unique(se$donor) %>% gsub('BRx', 'Brx-', .)
```
<!-- Write DGE results to xlsx file -->
```{r st1-write-xlsx}
# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'
file_xlsx <- file.path('./docs/file',rmd_file, 'st_1_dge.xlsx')

dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)

# Generate workbook
wb <- createWorkbook()
for(i in names(dge_list)) {
  use_dge <- dge_list[[i]]$lrt_time
  names(use_dge) <- (as.numeric(names(use_dge)) / 60) %>% 
    as.character
  for(j in names(use_dge)) {
    sheet_name <- paste0(i, ' timepoint ', j, 'h')
    addWorksheet(wb, sheet_name)
    res <- use_dge[[j]] %>% 
      dplyr::select(gene_name,
                    ensembl_gene_id,
                    gene_type,
                    log2FoldChange,
                    pvalue,
                    padj,
                    description) %>% 
      arrange(padj)
    writeData(wb, sheet_name, res)
  }
}
saveWorkbook(wb, file_xlsx, TRUE)

dge_file_xlsx <- file_xlsx
```

- [**Supplementary Table 1 : Bulk RNA-sequencing differential gene expression analysis**](`r gsub("docs/", "" , dge_file_xlsx)`). Table listing genes differentially expressed in BR16, Brx-07 and Brx-50 CTC cell lines incubated in static conditions (control) versus circulated through the microfluidic platform (shear stress) over time.

```{r st2-write-xlsx}
test_type <- 'GSEA'
gset_collection <- 'msigdb.c2.cp'
use_gse_omnibus <- time_gse_omnibus
names(use_gse_omnibus) <- names(use_gse_omnibus) %>% gsub('BRx', 'Brx-', .)

# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'
file_xlsx <- file.path('./docs/file',rmd_file, 'st_2_gsea.xlsx')

dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)

# Generate workbook
wb <- createWorkbook()

for(i in names(use_gse_omnibus)) {
  use_gse_i <- use_gse_omnibus[[i]]
  names(use_gse_i) <- (as.numeric(names(use_gse_i)) / 60) %>% 
    as.character
  for(j in names(use_gse_i)) {
    sheet_name <- paste0(i, ' timepoint ', j, 'h')
    addWorksheet(wb, sheet_name)
    res <- use_gse_i[[j]][[test_type]][[gset_collection]]@result %>%
      filter(grepl('REACTOME|KEGG|WP_', ID)) %>% 
      filter(!ID %in% 'KEGG_TRYPTOPHAN_METABOLISM') %>% 
      filter(!ID %in% 'WP_GLUCURONIDATION') %>% 
      filter(!ID %in% 'WP_MAPK_SIGNALING_PATHWAY') %>% 
      mutate(
        Description = clean_msigdb_names(Description)
      ) %>% 
      dplyr::select(ID,
                    Description,
                    setSize,
                    NES,
                    pvalue,
                    p.adjust,
                    leading_edge,
                    Count, 
                    GeneRatio,
                    core_enrichment) %>% 
      arrange(pvalue)
    writeData(wb, sheet_name, res)
  }  
  
}
saveWorkbook(wb, file_xlsx, TRUE)
gsea_file_xlsx <- file_xlsx
```

- [**Supplementary Table 2 : Bulk RNA-sequencing gene set enrichment analysis (GSEA)**](`r gsub("docs/", "" , gsea_file_xlsx)`). Table listing GSEA gene sets from differentially expressed genes in BR16, Brx-07 and Brx-50 CTC cell lines incubated in static conditions (control) versus circulated through the microfluidic platform (shear stress) over time. 



## Extended Data Figures 3a-c

```{r dge-deseq2-lrt-time-summary-conf}
p_thrs <- 0.01

dir_output <- dirname(params$output_dir)

dge_table_by_time <- foreach(use_model = unique(se$donor)) %do% {
  dge <- readRDS(file = file.path(dir_output, tolower(use_model), 'dge_deseq2.rds'))
  foreach(i = names(dge$lrt_time), .combine = rbind) %do% {
    data.frame(
      time = i,
      all = dge$lrt_time[[i]] %>% filter(padj < p_thrs)  %>% nrow,
      up =  dge$lrt_time[[i]] %>% filter(padj < p_thrs & log2FoldChange > 0) %>% nrow,
      down =  dge$lrt_time[[i]] %>% filter(padj < p_thrs & log2FoldChange < 0) %>% nrow
    )
  }
}
names(dge_table_by_time) <- tolower(unique(se$donor))
```

### Extended Data Figure 3a: Differentially expressed genes in BR16

Bar plot showing the number of differentially expressed genes in BR16 CTCs over time, comparing cells incubated in static conditions (control) and cells circulated through the microfluidic platform (shear stress) (bulk RNA-sequencing).

```{r dge-deseq2-lrt-time-summary-barplot-br16}
dge_table_by_time[['br16']] %>% 
  pivot_longer(cols = one_of(c('up', 'down')), names_to = 'set', values_to = 'genes') %>% 
  mutate(
    time = factor(time, levels = names(time_palette)),
    set = factor(set, levels = c('up', 'down'))) %>% 
  ggplot(aes(time, genes, fill = set)) +
  geom_col() +
  scale_fill_manual(values = c(up = brewer.pal(n = 11, name ="BrBG")[3], down = brewer.pal(n = 11, name ="BrBG")[9])) +
  labs(
    x = 'Time (minutes)',
    y = 'Number of gene',
    fill = '',
    title = 'Number of significant genes by time',
    caption = paste("Adjusted P value <", p_thrs)
  )
```


### Extended Data Figure 3b: Differentially expressed genes in Brx-07

Bar plot showing the number of differentially expressed genes in Brx-07 CTCs over time, comparing cells incubated in static conditions (control) and cells circulated through the microfluidic platform (shear stress) (bulk RNA-sequencing).


```{r dge-deseq2-lrt-time-summary-barplot-brx07}
dge_table_by_time[['brx07']] %>% 
  pivot_longer(cols = one_of(c('up', 'down')), names_to = 'set', values_to = 'genes') %>% 
  mutate(
    time = factor(time, levels = names(time_palette)),
    set = factor(set, levels = c('up', 'down'))) %>% 
  ggplot(aes(time, genes, fill = set)) +
  geom_col() +
  scale_fill_manual(values = c(up = brewer.pal(n = 11, name ="BrBG")[3], down = brewer.pal(n = 11, name ="BrBG")[9])) +
  labs(
    x = 'Time (minutes)',
    y = 'Number of gene',
    fill = '',
    title = 'Number of significant genes by time',
    caption = paste("Adjusted P value <", p_thrs)
  )
```


### Extended Data Figure 3c: Differentially expressed genes in Brx-50

Bar plot showing the number of differentially expressed genes in Brx-50 CTCs over time, comparing cells incubated in static conditions (control) and cells circulated through the microfluidic platform (shear stress) (bulk RNA-sequencing).


```{r dge-deseq2-lrt-time-summary-barplot-brx50}
dge_table_by_time[['brx50']] %>% 
  pivot_longer(cols = one_of(c('up', 'down')), names_to = 'set', values_to = 'genes') %>% 
  mutate(
    time = factor(time, levels = names(time_palette)),
    set = factor(set, levels = c('up', 'down'))) %>% 
  ggplot(aes(time, genes, fill = set)) +
  geom_col() +
  scale_fill_manual(values = c(up = brewer.pal(n = 11, name ="BrBG")[3], down = brewer.pal(n = 11, name ="BrBG")[9])) +
  labs(
    x = 'Time (minutes)',
    y = 'Number of gene',
    fill = '',
    title = 'Number of significant genes by time',
    caption = paste("Adjusted P value <", p_thrs)
  )
```


## Figure 1b: Heatmap of differential expressed genes

Heatmap showing row-normalized expression levels of differentially expressed genes (adjusted P value < 0.01, n = 6’982) for indicated CTC cell lines, between samples incubated in static conditions (control) and samples circulated through the microfluidic platform (shear stress) for varying durations. Annotated genes represent cytochrome P450 and AHR pathway related genes associated with xenobiotic response elements. 

<!-- Main heatmap -->
```{r dge-deseq2-lrt-summary-heatmap-plot, fig.width = 2, fig.asp = 0.88}
zmat <- zmat_topgenes

# Genes to annotate
use_genes_annot <- c('CYP1A1', 'CYP1B1', 'TIPARP', 
                     'CYP1A2', 
                     'NQO1', 'GSTA2', 'UGT1A1', 'UGT1A6', 'NFE2L2',
                     'AHR', 'LRAT', 'IL22',  'LHX4' )

# Z-score, using mean in Control_0 as scaling factor
use_cols <- colnames(zmat) %in% colnames(se)[se$group != 'Control_0']
use_zmat <- zmat[,use_cols]
use_cols_ord <- colData(se[,colnames(use_zmat)]) %>% data.frame %>% 
  arrange(condition, time, donor) %>% pull(sample_alias)
use_zmat <- use_zmat[,use_cols_ord]


# Select only samples from Chip
chip_samples <- colData(se[,colnames(use_zmat)]) %>% 
  data.frame %>% 
  filter(condition == 'Chip') %>% 
  pull(sample_alias)
use_zmat <- use_zmat[,chip_samples]



# Set heatmap color limiting the range matrix
zlim <- c(-2, 2)
zmax <- range(use_zmat) %>% abs %>% max()
col_fun <-  colorRamp2(
  c(-zmax, seq(zlim[1], zlim[2], length.out = 7), zmax),
  rev(brewer.pal(n = 9, name ="RdBu")))

# Mark annotation
rownames(use_zmat) <- rowData(se[rownames(use_zmat),])$gene_name
use_genes_annot <- intersect(use_genes_annot, rownames(use_zmat))
row_mark_anno <-  rowAnnotation(
  foo = anno_mark(at = match(use_genes_annot, rownames(use_zmat)), 
                  labels = use_genes_annot,
                  labels_gp = gpar(fontsize = 3),
                  lines_gp = gpar(lwd = 0.5),
                  link_width = unit(3, "mm"))
  )



# Column annotation
column_ha <- HeatmapAnnotation(
  model = colData(se[,colnames(use_zmat)])$donor,
  col = list(model = alpha(donor_palette, 0.8)),
  # height = unit(5, "mm"),
  simple_anno_size = unit(1, "mm"),
  show_annotation_name = FALSE,
  show_legend = FALSE
    # annotation_legend_param = list(title_gp = gpar(fontsize = 5),
  #                                labels_gp = gpar(fontsize = 5),
  #                                grid_width = unit(0.15, "cm"),
  #                                title = '')
  )



# Print the final heatmap
Heatmap(
  use_zmat, 
  name = 'z\nscore',
  col =  col_fun,
  
  right_annotation = row_mark_anno,
  top_annotation = column_ha,
  
  cluster_columns = FALSE,
  show_row_names = FALSE, 
  show_column_names = FALSE,
  row_title = NULL, # remove to see if row split names are equal to those added at left_annotation
  row_dend_reorder = TRUE,
  show_row_dend = FALSE,
  # row_split = row_split,
  column_split = se[,colnames(use_zmat)]$time,
  column_gap = unit(0.5, "mm"),
  border = TRUE,
  border_gp = gpar(col = "black", lwd = 0.5),
  
  column_names_gp = gpar(fontsize = 6),
  # heatmap_legend_param = list(title_gp = gpar(fontsize = 5), 
  #                             labels_gp = gpar(fontsize = 5),
  #                             grid_width = unit(0.15, "cm")),
  column_title_gp = gpar(fontsize = 6),
  
  use_raster = TRUE,
  show_heatmap_legend = FALSE
  )
```

<!-- Legend -->
```{r dge-deseq2-lrt-summary-heatmap-plot-legends-v, fig.width = 2, fig.asp = 0.88}
ht_legend <-  Legend(col_fun = col_fun,
              title = "z\nscore",
              title_gp = gpar(fontsize = 5),
              labels_gp = gpar(fontsize = 5)
              )

column_ha_legend <-  Legend(labels = names(donor_palette), 
              legend_gp = gpar(fill = alpha(donor_palette, 0.8)), 
              title = "",
              # nrow = 1,
              title_gp = gpar(fontsize = 5),
              labels_gp = gpar(fontsize = 5)
              # grid_width = unit(0.15, "cm")
              )

pd <-  packLegend(ht_legend, column_ha_legend, max_width = unit(10, "cm"), 
                  row_gap = unit(5, "mm"))
draw(pd)
```


## Figure 1d: Time-series of log2FC

Fold-change of cytochrome (CYP) P450 pathway related genes (upper), and AHR pathway related genes (lower) over time in samples upon incubation in static conditions (control) and samples circulated through the microfluidic platform (shear stress) across the indicated CTC cell lines. 

### CYP-450 genes
```{r gene_candidates-fc-time-cyp-plot, fig.width=4.98, fig.asp=0.22}
use_linewidth <- 1/2.141959
# use_genes_annot <- c('CYP1A1', 'CYP1B1', 'TIPARP', 
#                      'CYP1A2', 
#                      'NQO1', 'GSTA2', 'UGT1A1', 'UGT1A6', 'NFE2L2',
#                      	'AHR', 'LRAT', 'IL22',  'LHX4' )

use_genes_annot <- c('CYP1A1', 'CYP1B1')
use_palette <- get_scater_palette('tableau20')[seq_along(use_genes_annot)] %>% 
  set_names(use_genes_annot)

x <- dge_df %>% 
  filter(gene_name %in% use_genes_annot) %>% 
  mutate(
    time = as.numeric(time),
    padj = ifelse(is.na(padj), 1, padj),
    padj = ifelse(padj < 1e-10, 1e-10, padj),
    log_padj = -log10(padj)
  )


x_breaks <- x$time %>% as.character %>% as.numeric %>% sort %>% unique
y_range <- x$log2FoldChange %>% range
y_lim <- c(floor(y_range[1]), ceiling(y_range[2]))

# Add time 0
add_time0 <- x %>% 
  group_by(donor, gene_name) %>% 
  summarise(time = 0) %>% 
  mutate(
    log2FoldChange = 0,
    log_padj = 0
  )

x <- rbind(
  x %>% 
    select(donor, gene_name, time, log2FoldChange, log_padj),
  add_time0
  )


donor_levels <- x$donor %>% unique()
plots_list <- foreach(use_donor = donor_levels) %do% {
  x_ends <- x %>% filter(time == 480 & donor == use_donor)
  
  x %>% 
    filter(donor == use_donor) %>%
    ggplot(aes(time, log2FoldChange, color = gene_name,
               group = gene_name, size = log_padj)) +
    geom_point() +
    geom_line(linewidth = use_linewidth) +
    geom_hline(
      yintercept = 0, 
      linetype = 'dashed', 
      color = 'grey', 
      linewidth = use_linewidth) +
    scale_color_manual(values = use_palette) +
    scale_size_continuous(
      range = c(0.2, 1),
      limits = x$log_padj %>% range
      ) +
    scale_x_continuous(
        expand = expansion(mult = c(0, 0.05)),
        breaks = c(0, x_breaks)
        ) + 
    scale_y_continuous(
      expand = expansion(mult = c(0.01, 0)),
      limits = y_lim,
      sec.axis = dup_axis(
        breaks = x_ends$log2FoldChange,
        labels = x_ends$gene_name,
        name = NULL)
    ) +
    theme(
      axis.line.y.right = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.text.y.right = element_text(size = 4, margin = margin(0, 0, 0, 0)),
      
      axis.ticks.y.left = element_line(color = "black", linewidth = use_linewidth),
      axis.ticks.x = element_line(color = "black", linewidth = use_linewidth),
      axis.line.x = element_line(color = "black", linewidth = use_linewidth),
      axis.line.y.left = element_line(color = "black", linewidth = use_linewidth),
      axis.text.x = element_text(angle = 45, hjust = 1),
      # axis.text = element_text(size = 6),
      # legend.text = element_text(size = 6), 
      # legend.title = element_text(size = 6),
      
      # plot.margin = margin(14, 7, 3, 1.5),
      plot.margin = margin(0, 0, 0, 0),
      
    ) +
    guides(color = 'none') +
    labs(
        x = 'Time (minutes)',
        size = expression(paste("lo", g[10],"(FDR)")),
        title = use_donor
      ) +
    ylab(NULL)
}

# Shared legend: extract the legend from one of the plots
legend <- get_legend(
  # create some space to the left of the legend
  plots_list[[1]] + theme(legend.box.margin = margin(0, 0, 0, 12))
)

# Remove legends from plots
plots_list <- foreach(p = plots_list) %do% {
  p + theme(legend.position = "none")
}

# Add y lab to first plot
plots_list[[1]] <- plots_list[[1]] + ylab(expression(paste("lo", g[2],"(Fold change)")))

# Generate drid of plots
prow <- plot_grid(plotlist = plots_list, 
          align = 'vh',
          hjust = -1,
          nrow = 1)

# Generate final
plot_grid(prow, legend, rel_widths = c(3, .4))

# Generate final without legend
plot_grid(plotlist = plots_list, 
          align = 'vh',
          hjust = -1,
          nrow = 1)
```

### AhR targeting genes genes
```{r gene_candidates-fc-time-ahr-plot, fig.width=4.98, fig.asp=0.22}
use_linewidth <- 1/2.141959
use_genes_annot <- c('TIPARP',
                     'NQO1', 'GSTA2', 'UGT1A1', 'UGT1A6', 'NFE2L2',
                     	'AHR', 'LRAT', 'IL22',  'LHX4' )

use_palette <- get_scater_palette('tableau20')[seq_along(use_genes_annot)] %>% 
  set_names(use_genes_annot)

x <- dge_df %>% 
  filter(gene_name %in% use_genes_annot) %>% 
  mutate(
    time = as.numeric(time),
    padj = ifelse(is.na(padj), 1, padj),
    padj = ifelse(padj < 1e-10, 1e-10, padj),
    log_padj = -log10(padj)
  )


x_breaks <- x$time %>% as.character %>% as.numeric %>% sort %>% unique
y_range <- x$log2FoldChange %>% range
y_lim <- c(floor(y_range[1]), ceiling(y_range[2]))

# Add time 0
add_time0 <- x %>% 
  group_by(donor, gene_name) %>% 
  summarise(time = 0) %>% 
  mutate(
    log2FoldChange = 0,
    log_padj = 0
  )

x <- rbind(
  x %>% 
    select(donor, gene_name, time, log2FoldChange, log_padj),
  add_time0
  )


donor_levels <- x$donor %>% unique()
plots_list <- foreach(use_donor = donor_levels) %do% {
  x_ends <- x %>% filter(time == 480 & donor == use_donor)
  
  x %>% 
    filter(donor == use_donor) %>%
    ggplot(aes(time, log2FoldChange, color = gene_name,
               group = gene_name, size = log_padj)) +
    geom_point() +
    geom_line(linewidth = use_linewidth) +
    geom_hline(
      yintercept = 0, 
      linetype = 'dashed', 
      color = 'grey', 
      linewidth = use_linewidth) +
    scale_color_manual(values = use_palette) +
    scale_size_continuous(
      range = c(0.2, 1),
      limits = x$log_padj %>% range
      ) +
    scale_x_continuous(
        expand = expansion(mult = c(0, 0.05)),
        breaks = c(0, x_breaks)
        ) + 
    scale_y_continuous(
      expand = expansion(mult = c(0.01, 0)),
      limits = y_lim,
      sec.axis = dup_axis(
        breaks = x_ends$log2FoldChange,
        labels = x_ends$gene_name,
        name = NULL)
    ) +
    theme(
      axis.line.y.right = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.text.y.right = element_text(size = 4, margin = margin(0, 0, 0, 0)),
      
      axis.ticks.y.left = element_line(color = "black", linewidth = use_linewidth),
      axis.ticks.x = element_line(color = "black", linewidth = use_linewidth),
      axis.line.x = element_line(color = "black", linewidth = use_linewidth),
      axis.line.y.left = element_line(color = "black", linewidth = use_linewidth),
      axis.text.x = element_text(angle = 45, hjust = 1),
      # axis.text = element_text(size = 6),
      # legend.text = element_text(size = 6), 
      # legend.title = element_text(size = 6),
      
      # plot.margin = margin(14, 7, 3, 1.5),
      plot.margin = margin(0, 0, 0, 0),
      
    ) +
    guides(color = 'none') +
    labs(
        x = 'Time (minutes)',
        size = expression(paste("lo", g[10],"(FDR)")),
        title = use_donor
      ) +
    ylab(NULL)
}

# Shared legend: extract the legend from one of the plots
legend <- get_legend(
  # create some space to the left of the legend
  plots_list[[1]] + theme(legend.box.margin = margin(0, 0, 0, 12))
)

# Remove legends from plots
plots_list <- foreach(p = plots_list) %do% {
  p + theme(legend.position = "none")
}

# Add y lab to first plot
plots_list[[1]] <- plots_list[[1]] + ylab(expression(paste("lo", g[2],"(Fold change)")))

# Generate drid of plots
prow <- plot_grid(plotlist = plots_list, 
          align = 'vh',
          hjust = -1,
          nrow = 1)

# Generate final
plot_grid(prow, legend, rel_widths = c(3, .4))

```


## Figure 2c and Extended Data Figure 3d: Functional Enrichment analysis
```{r gse-omnibus-time-c2cp-conf}
test_type <- 'GSEA'
gset_collection <- 'msigdb.c2.cp'
p_thrs <- 0.0001
n_split <- 10


# Combine all results, selecting top_n gene sets per analysis
comb_gse <- foreach(donor = names(time_gse_omnibus), .combine = rbind) %do% {
  foreach(tm = names(time_gse_omnibus[[donor]]), .combine = rbind) %do% {
    time_gse_omnibus[[donor]][[tm]][[test_type]][[gset_collection]]@result %>% 
      mutate(time = tm)
  } %>% 
    mutate(donor = donor)
} %>% 
  mutate(time = factor(time, levels = levels(se$time)))

# Keep only reactome and kegg and WP, removing overlapping pathways
comb_gse %<>% 
  filter(grepl('REACTOME|KEGG|WP_', ID)) %>% 
  filter(!ID %in% 'KEGG_TRYPTOPHAN_METABOLISM') %>% 
  filter(!ID %in% 'WP_GLUCURONIDATION') %>% 
  filter(!ID %in% 'WP_MAPK_SIGNALING_PATHWAY') %>% 
  mutate(
    Description = clean_msigdb_names(Description),
    ID = clean_msigdb_names(ID)
  )

# Generate similarity matrix:
  # We combine all significant gene sets for each timepoint. To do so, we remove the duplicated gene sets by selecting the one with the lowest adjusted P value. Then we update the list of core genes by combining all the genes from all the comparisons. Finally, we run the enrichplot::pairwise_termsim function to calculate the similarity matrix.
fake_gseaResult <-  time_gse_omnibus[[1]][[1]][[test_type]][[gset_collection]]
# Select comparisons with the lowest adjusted P value
top_by_time_pthrs <-  comb_gse %>% 
  filter(p.adjust < p_thrs) %>% 
  arrange(p.adjust) %>% 
  filter(!duplicated(ID))
rownames(top_by_time_pthrs) <- NULL
top_by_time_pthrs %<>% mutate(torow = ID) %>% column_to_rownames('torow')
# Update the list of core genes by combining all the genes from all the comparisons.
combine_core_enrichment <- foreach(i = top_by_time_pthrs$ID) %do% {
  comb_gse %>% filter(ID == i) %>% 
    pull(core_enrichment) %>% 
    paste0(., collapse = '/') %>% 
    strsplit(., '/') %>% 
    unlist %>% 
    unique %>% paste(.,collapse = '/')
} %>% unlist()
names(combine_core_enrichment) <- top_by_time_pthrs$ID
top_by_time_pthrs$core_enrichment <- combine_core_enrichment[ top_by_time_pthrs$ID]
fake_gseaResult@result <- top_by_time_pthrs
#Generate similarity matrix
fake_gseaResult <- pairwise_termsim(fake_gseaResult)

similarity_matrix <- fake_gseaResult@termsim
# Transform matrix to symmetric
use_mat <- similarity_matrix
for(x in rownames(use_mat)){
  for(y in colnames(use_mat)) {
    if(x == y) {
       use_mat[x,y] <- 1
    } else {
      max_sim <- max(c(use_mat[x,y], use_mat[y,x]), na.rm = TRUE)
      use_mat[x,y] <- max_sim
      use_mat[y,x] <- max_sim
    }
  }
}
rownames(use_mat) <- clean_msigdb_names(rownames(use_mat))
colnames(use_mat) <- clean_msigdb_names(colnames(use_mat))

# Cluster the similarity matrix
# Use the ward.D method to avoid overlapping ancestor nodes of each group
hc <- hclust(
  as.dist(1- use_mat),
  method = 'ward.D'
)
split <- cutree(hc, k=n_split) %>% data.frame %>% set_names('cluster')
split$curated_id <- clean_msigdb_names(rownames(split))

# We add cluster number 11. Based on GSEa results, clsuter 10 have two different groups of gene sets (Up and Down)
split %<>% 
  mutate(
    cluster = ifelse(
      curated_id %in% c('SYNTHESIS OF DNA', 'ORC1 REMOVAL FROM CHROMATIN', 'ANTIGEN PROCESSING CROSS PRESENTATION'),
      11,
      cluster
    )
  )

# Annotate clusters.  For each cluster, sort the genesets by GeneRatio and P.adjust, and select the top ones
fake_gseaResult@result %<>% 
  left_join(split %>% rownames_to_column('ID'))
# foreach(i = unique(split$cluster), .combine = rbind) %do% {
#   top_generatio <- fake_gseaResult@result %>%
#     filter(cluster == i) %>%
#     arrange(desc(GeneRatio)) %>%
#     head(1) %>%
#     pull(ID) %>%
#     clean_msigdb_names
#   top_padj <- fake_gseaResult@result %>%
#     filter(cluster == i) %>%
#     arrange(p.adjust) %>%
#     head(1) %>%
#     pull(ID) %>%
#     clean_msigdb_names
#   c(cluster = i, GeneRatio = top_generatio, p.adjust = top_padj)
# }

# Manually annotate clusters, using top gene sets per cluster as guide
# fake_gseaResult@result %>% filter(cluster == '1') %>% select(ID, NES, p.adjust, GeneRatio, core_enrichment)
split$cluster_name <- case_when(
  split$cluster == 1 ~ 'NGF-stimulated transcription',
  split$cluster == 2 ~ 'AhR pathway and Oxidative stress response',
  split$cluster == 3 ~ 'Inflammation / Cytokines signaling',
  split$cluster == 4 ~ 'Metabolism of xenobiotics by Cytochrome P450 / Glucuronidation',
  split$cluster == 5 ~ 'Interferon signaling',
  split$cluster == 6 ~ 'Cytochrome P450-Catalyzed Oxidations',
  split$cluster == 7 ~ 'Translation',
  split$cluster == 8 ~ 'Heat shock response',
  split$cluster == 9 ~ 'HSF1 activation',
  split$cluster == 10 ~ 'KEAP1-NFE2L2 pathway',
  split$cluster == 11 ~ 'Other',
  TRUE ~ 'Other'
)
split %<>% 
  mutate(
    cluster_fct = factor(cluster, levels = as.character(1:n_split)),
    cluster_name =  fct_reorder(cluster_name, cluster)
  )

relevel_clusters <- c(
  'AhR pathway and Oxidative stress response',
  'Metabolism of xenobiotics by Cytochrome P450 / Glucuronidation',
  'Cytochrome P450-Catalyzed Oxidations',
  'HSF1 activation',
  'Heat shock response',
  'KEAP1-NFE2L2 pathway',
  'Inflammation / Cytokines signaling',
  'NGF-stimulated transcription',
  'Translation',
  'Interferon signaling',
  'Other'
)




# We select significant gene sets in all test (donor, time)
top_all <- comb_gse %>% 
  filter(p.adjust < p_thrs) %>% 
  pull(ID) %>% unique

res_top <- comb_gse %>% 
  filter(ID %in% top_all) %>% 
  left_join(split, by = c('ID' = 'curated_id'))
```

### Extended Data Figure 2d

Gene set enrichment analysis (GSEA) plots for representative gene sets in BR16, Brx-07 and Brx-50 CTC lines across circulation time points. NES, normalized enrichment score; FDR, false discovery rate.

```{r gse-omnibus-time-dotplot-c2cp-time-plot-full, fig.width = 3, fig.asp = 1.10}
use_linewidth <- 1/2.141959

res_top %>% 
  mutate(
    p.sig = p.adjust < 0.05,
    p.adjust = -log10( p.adjust),
    cluster_name = factor(cluster_name, levels = relevel_clusters)
    ) %>% 
  ggplot(aes(time, Description, fill = NES, color = p.sig, size = p.adjust)) +
  geom_point(shape = 21, stroke = 0.1) +
  facet_grid(cols = vars(donor), rows = vars(cluster_name),
             scale = 'free', space = 'free',
             labeller = label_wrap_gen(20) ) +
  scale_fill_distiller(palette = "RdBu", na.value = "grey70") +
  scale_color_manual(values = c('white', 'black')) +
  scale_y_discrete(labels = default_labeller(50)) +
  scale_size_continuous(range=c(0.1, 0.9)) +
  labs(y = '', 
       x = '',  
       size = expression(paste("lo", g[10],"(FDR)")),
       color = 'FDR < 0.05'
       ) +
  theme(
    text = element_text(size = 4),
    axis.line = element_blank(),
    # axis.ticks.y = element_blank(),
    axis.ticks.y = element_line(linewidth = use_linewidth/2),
    axis.ticks.x = element_line(linewidth = use_linewidth/2),
    panel.border = element_rect(color = "black", fill = NA, linewidth = use_linewidth),
   
    strip.background = element_blank(),
    axis.text.y = element_text(size = 2),
    axis.text.x = element_text(size = 3, angle = 45, hjust = 1),
    legend.text = element_text(size = 2),
    legend.title = element_text(size = 2),
    legend.key.width = unit(0.1, "cm"),
    
    strip.text.y = element_text(size = 2, angle = 0, hjust = 0),
    panel.spacing = unit(0.07, "lines")
  )
```


### Figure 2c

Gene set enrichment analysis (GSEA) plots for representative gene sets in indicated CTC cell lines across time points of circulation through the microfluidic platform (shear stress) compared to incubation in static conditions (control). NES, normalized enrichment score; FDR, false discovery rate.

```{r gse-omnibus-time-dotplot-c2cp-time-plot-representative, fig.width = 2, fig.asp = 0.90}
use_linewidth <- 1/2.141959

selected_gene_sets <- c(
  'AhR pathway' = '2873 ARYL HYDROCARBON RECEPTOR PATHWAY',
  'Oxidative stress' = 'OXIDATIVE STRESS RESPONSE',
  'Metabolism of xenobiotics by Cytochrome P450' = "METABOLISM OF XENOBIOTICS BY CYTOCHROME P450",
  'Glucuronidation' = 'GLUCURONIDATION',
  'Cytochrome P450-Catalyzed Oxidations' = "OXIDATION BY CYTOCHROME P450",
  'HSF1 activation' = 'HSF1 ACTIVATION',
  'Heat shock response' = "CELLULAR RESPONSE TO HEAT STRESS",
  'KEAP1-NFE2L2 pathway' = "KEAP1 NFE2L2 PATHWAY",
  'Cytokines signaling' = "TGFBETA SIGNALING PATHWAY",
  'NGF-stimulated transcription' = "NGF STIMULATED TRANSCRIPTION",
  'Translation' = "EUKARYOTIC TRANSLATION INITIATION",
  'Interferon signaling' = "INTERFERON SIGNALING"
) %>% data.frame %>% set_names('Description') %>% rownames_to_column('group_name')

res_top %>% 
  left_join(selected_gene_sets, by = 'Description') %>%
  filter(!is.na(group_name)) %>% 
  mutate(
    p.sig = p.adjust < 0.05,
    p.adjust = -log10( p.adjust),
    group_name = factor(group_name, levels = rev(selected_gene_sets$group_name))
    ) %>% 
  ggplot(aes(time, group_name, fill = NES, color = p.sig, size = p.adjust)) +
  geom_point(shape = 21, stroke = 0.1) +
  facet_grid(cols = vars(donor),
             scale = 'free', space = 'free') +
  scale_fill_distiller(palette = "RdBu", na.value = "grey70") +
  scale_color_manual(values = c('white', 'black')) +
  scale_y_discrete(labels = default_labeller(30)) +
  scale_size_continuous(range=c(0.7, 2)) +
  labs(y = '', 
       x = 'Time (minutes)',  
       size = expression(paste("lo", g[10],"(FDR)")),
       color = 'FDR < 0.05'
       ) +
  theme(
    text = element_text(size = 4),
    axis.line = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_line(linewidth = use_linewidth/2),
    panel.border = element_rect(color = "black", fill = NA, linewidth = use_linewidth),
   
    strip.background = element_blank(),
    # axis.text.y = element_blank(),
    axis.text.y = element_text(size = 4),
    axis.text.x = element_text(size = 4, angle = 90, vjust = 0.5, hjust = 1),
    legend.text = element_text(size = 2),
    legend.title = element_text(size = 2),
    legend.key.width = unit(0.1, "cm"),
    
    panel.spacing = unit(0.04, "lines")
  )
```


