---
title: "Analysis of gene KO and shear stress in BR16"
subtitle: "Differential gene expression at gene level"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  output_dir: ./output/rnaseq/bulk-gene_ko/br16/differential_expression/gene
  se_file: ./data/rnaseq/bulk-ko/br16/se.rds
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(knitr)
library(ggplot2)
library(tibble)
library(tidyr)
library(readr)
library(purrr)
library(dplyr)
library(forcats)
library(kableExtra)
library(RColorBrewer)
library(ggrepel)
library(openxlsx)

library(showtext)
library(foreach)
library(magrittr)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(ComplexHeatmap)
library(circlize)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/subchunkify.r')
source('./code/R-functions/pairwise_terms_jc.r')
clean_msigdb_names <- function(x) x %>% gsub('REACTOME_', '', .) %>% gsub('WP_', '', .) %>% gsub('BIOCARTA_', '', .) %>% gsub('KEGG_', '', .) %>% gsub('PID_', '', .) %>% gsub('GOBP_', '', .) %>% gsub('_', ' ', .) %>% str_to_sentence

```


Load MSigDB gene sets
```{r load-msigdb}
msigdb_vs <- 'v2022.1.Hs'
gmt_files_symbols <- list(
  msigdb.c2.cp = paste0('./data/resources/MSigDB/', msigdb_vs, '/c2.cp.', msigdb_vs, '.symbols.gmt')
)
```

## Data processing

Load data
```{r load-data}
se_files <- strsplit(params$se_file, ",") %>% unlist

if(length(se_files) > 1) {
  se_list <- foreach(se_file = se_files) %do% {
    readRDS(se_file)
  }
  se <- merge_SummarizedExperiment(se_list)
} else {
  se <- readRDS(se_files)
}
```

Modify colData and rowData
```{r dw-mod-data}
colData(se) %<>% data.frame %>% 
  mutate(
    ko = ifelse(gene_ko == 'WT', 'wt', 'ko') %>% 
      factor(levels = c('wt', 'ko')),
  ) %>% 
  DataFrame

rownames(se) <- rowData(se)$gene_name %>% make.names(unique = TRUE)
```


## Define of comparisons

Define groups of analysis
```{r conf-comparisons}
x <- colData(se) %>% data.frame
use_subsets <- setdiff(unique(x$gene_ko), 'WT')
subset_list <- foreach(i = use_subsets) %do% {
  x %>% 
    filter(gene_ko %in% c('WT', i)) %>% 
    rownames
}
names(subset_list) <- use_subsets
```

## Data generation

### Run differential expression analysis

Get Ensembl gene info
```{r de-ensembl, eval = FALSE}
ensembl <-  biomaRt::useEnsembl(biomart="genes", dataset="hsapiens_gene_ensembl",
                                mirror = 'useast')
gene_desc <- biomaRt::getBM(attributes=c('external_gene_name','description'),
                              filters = 'external_gene_name',
                              values = rowData(se)$gene_name, mart =ensembl) %>%
    dplyr::rename('gene_name' = 'external_gene_name') %>%
    unique

gene_desc <- gene_desc[!duplicated(gene_desc$gene_name),]
saveRDS(gene_desc, file = file.path(params$output_dir, 'ensembl_gene_desc.rds'))
```

```{r de-ensembl-load, eval = FALSE}
gene_desc <- readRDS(file = file.path(params$output_dir, 'ensembl_gene_desc.rds'))
```

Run differential expression analysis using DESeq for each subset
```{r deseq2-run, eval = FALSE}
use_subset <- subset_list[[1]]
dge_list <- foreach(use_subset = subset_list) %do%{
  use_se <- se[,use_subset]
  group_levels <- outer(levels(use_se$condition), levels(use_se$ko), paste, sep="_") %>% 
    as.vector
  use_se$group <- paste(use_se$condition, use_se$ko, sep = '_') %>% 
    factor(levels = group_levels)

  # Generate DESeq for group
  dds <- DESeqDataSet(use_se, design = ~ group)
  dds <- DESeq(dds)
   # resultsNames(dds)
  # Contrast: a character vector with exactly three elements: the name of a factor in the design formula, the name of the numerator level for the fold change, and the name of the denominator level for the fold change (simplest case)
  res_chip_in_wt <- results(dds, contrast = c("group", "Chip_wt", "Control_wt")) %>% 
    data.frame %>% 
    rownames_to_column('gene_name') %>% 
    left_join(
      gene_desc, by = 'gene_name'
    )
  res_chip_in_ko <- results(dds, contrast = c("group", "Chip_ko", "Control_ko")) %>% 
    data.frame %>% 
    rownames_to_column('gene_name') %>% 
    left_join(
      gene_desc, by = 'gene_name'
    )
  res_ko_in_control <- results(dds, contrast = c("group", "Control_ko", "Control_wt")) %>% 
    data.frame %>% 
    rownames_to_column('gene_name') %>% 
    left_join(
      gene_desc, by = 'gene_name'
    )
  res_ko_in_chip <- results(dds, contrast = c("group", "Chip_ko", "Chip_wt")) %>% 
    data.frame %>% 
    rownames_to_column('gene_name') %>% 
    left_join(
      gene_desc, by = 'gene_name'
    )
   
  
  # Generate DESeq for interaction
  dds <- DESeqDataSet(use_se, design = ~ condition + ko + condition:ko)
  dds <- DESeq(dds)
  # resultsNames(dds)
  # res_chip_wt_i <- results(dds, contrast = c("condition", "Control", "Chip"))
  # res_ko_in_control_i <- results(dds, contrast = c("ko", "wt", "ko"))
  res_i <- results(dds, name ="conditionChip.koko")  %>% 
    data.frame %>% 
    rownames_to_column('gene_name') %>% 
    left_join(
      gene_desc, by = 'gene_name'
    )
  # res_chip_wt== res_chip_wt_i
  # res_ko_in_control == res_ko_in_control_i
  
  # Interpretation of log2FC of interaction:
  # the log2FC of interaction (conditionChip.koko) represents the 
  # additional effect of ko in Chip condition compared to Control condition. 
  # log2FC(i) = log2FC(Chip_ko-over-Chip_wt) - log2FC(Control_ko-over-Control_wt)
  
  # Combine results in a single data.frame
  res_comb <- full_join(
    res_chip_in_wt,
    res_chip_in_ko,
    by = c('gene_name', 'description'), 
    suffix = c('_wt', '_ko')
  ) %>% 
    left_join(
      res_i %>% 
        dplyr::rename(
          log2FoldChange_int = log2FoldChange, 
          pvalue_int = pvalue, 
          padj_int = padj) %>% 
        dplyr::select(gene_name, ends_with('int'))
      ) %>% 
    dplyr::select(-description)%>% 
    left_join(
      gene_desc, by = 'gene_name'
    )
    
  
  # Return results
  res <- list(
    chip_in_wt = res_chip_in_wt,
    chip_in_ko = res_chip_in_ko,
    ko_in_control = res_ko_in_control,
    ko_in_chip = res_ko_in_chip,
    interaction = res_i,
    combined = res_comb
  )
  
  return(res)
}

names(dge_list) <- names(subset_list)
saveRDS(dge_list, file = file.path(params$output_dir, 'dge-deseq2.rds'))

```

### Run functional enrichment
```{r gse-run, eval = FALSE}
dge_list <- readRDS(file.path(params$output_dir, 'dge-deseq2.rds'))
dge_subset <-  dge_list[[2]]

gse_list <- foreach(dge_subset = dge_list) %do%{
  dge_subset$combined <- NULL
  gse_res_list <- foreach(dge = dge_subset) %do% {
    gse_res <- gse_omnibus(
      feature_names = dge$gene_name,
      p = dge$padj,
      fc = dge$log2FoldChange,
      fc_thrs = 0.2,
      gmt_files = gmt_files_symbols,
      
      run_enricher = TRUE,
      run_GSEA = TRUE,
      
      go_simplify_p = 0.05,
      args_gse = list(minGSSize = 10, maxGSSize = 250, 
                      pvalueCutoff = 1, 
                      BPPARAM = BiocParallel::MulticoreParam(1)),
      args_enricher = list(minGSSize = 10, maxGSSize = 250, 
                      pvalueCutoff = 1)
    )
    
    return(gse_res)
  }
  names(gse_res_list) <- names(dge_subset)
  return(gse_res_list)
}
names(gse_list) <- names(dge_list)

# Save objects
saveRDS(gse_list, file = file.path(params$output_dir,'gse-deseq2.rds'))

```

### Load data
```{r dge-gse-load}
dge_subset_list <- readRDS(file = file.path(params$output_dir, 'dge-deseq2.rds'))
gse_subset_list <- readRDS(file = file.path(params$output_dir, 'gse-deseq2.rds'))
```

## AHR
```{r AHR-inhibitor-conf}
use_subset <- 'AHR'
use_gse_list <- dge_subset_list[[use_subset]]
p_thrs <- 0.05
lfc_thrs <- 1
p_limit <- 1e-10
comparisons_indv <- setdiff(names(use_gse_list), 'combined')
candidate_genes <- c('CYP1A1', 'CYP1B1', 'NFE2L2', 'TIPARP', 'LHX4', 'CYP1A2', 'NQO1', 'AHR', 'UGT1A6', 'UGT1A1', 'CDC42')

# Classify genes by regulation
use_gse_list_update <- foreach(i = comparisons_indv) %do% {
  use_gse_list[[i]] %>%
    filter(!is.na(padj)) %>%
    mutate(
      regulation = case_when(
        padj <= p_thrs & log2FoldChange > lfc_thrs ~ 'up',
        padj <= p_thrs & log2FoldChange < -lfc_thrs ~ 'down',
        TRUE ~ 'ns'
      )
    )
}
names(use_gse_list_update) <- comparisons_indv
use_gse_list_update[['combined']] <- use_gse_list[['combined']]
use_gse_list <- use_gse_list_update
rm(use_gse_list_update)

# Prepare variables to run template
dge_list <- use_gse_list
gse_list <- gse_subset_list[[use_subset]]
file_prefix <- use_subset
```


### Differential gene expression
```{r AHR-inhibitor-knitr-template-deseq, eval = TRUE}
# # knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-deg-ko-deseq.Rmd")
  )
```
`r knit(text = src)`

### Functional enrichment analysis
```{r AHR-inhibitor-knitr-template-gse, eval = TRUE}
# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-gse-omnibus-drug_inhibitors.Rmd")
  )
```
`r knit(text = src)`
 


## CYP1A1 and CYP1B1
```{r CYP1A1_CYP1B1-inhibitor-conf}
use_subset <- 'CYP1A1_CYP1B1'
use_gse_list <- dge_subset_list[[use_subset]]
p_thrs <- 0.05
lfc_thrs <- 1
p_limit <- 1e-10
comparisons_indv <- setdiff(names(use_gse_list), 'combined')
candidate_genes <- c('CYP1A1', 'CYP1B1', 'NFE2L2', 'TIPARP', 'LHX4', 'CYP1A2', 'NQO1', 'AHR', 'UGT1A6', 'UGT1A1', 'CDC42')

# Classify genes by regulation
use_gse_list_update <- foreach(i = comparisons_indv) %do% {
  use_gse_list[[i]] %>%
    filter(!is.na(padj)) %>%
    mutate(
      regulation = case_when(
        padj <= p_thrs & log2FoldChange > lfc_thrs ~ 'up',
        padj <= p_thrs & log2FoldChange < -lfc_thrs ~ 'down',
        TRUE ~ 'ns'
      )
    )
}
names(use_gse_list_update) <- comparisons_indv
use_gse_list_update[['combined']] <- use_gse_list[['combined']]
use_gse_list <- use_gse_list_update
rm(use_gse_list_update)

# Prepare variables to run template
dge_list <- use_gse_list
gse_list <- gse_subset_list[[use_subset]]
file_prefix <- use_subset
```


### Differential gene expression
```{r CYP1A1_CYP1B1-inhibitor-knitr-template-deseq, eval = TRUE}
# # knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-deg-ko-deseq.Rmd")
  )
```
`r knit(text = src)`

### Functional enrichment analysis
```{r CYP1A1_CYP1B1-inhibitor-knitr-template-gse, eval = TRUE}
# knit template: https://workflowr.github.io/workflowr/articles/wflow-07-common-code.html
src <- knit_expand(
  file = here::here("analysis/templates/rnaseq-gse-omnibus-drug_inhibitors.Rmd")
  )
```
`r knit(text = src)`
 

